---
title: "Reproducible analysis for Zyck et al. 2026"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(dev = "cairo_png")
theme_set(theme_classic(base_family = "Liberation Sans"))
```


# Setup R and Environment
This scripts assumes the directory structure of the github repository and that all demultiplexed sequencing reads are in the `./analysis` directory

```{r}
library(OutFLANK)
library(bigsnpr)
library(plyr)
library(dplyr)
library(stringr)
library(reshape2)
library(tidyr)
library(ggplot2)
library(plotrix)
library(ggtext)
library(ggsurvfit)
library(patchwork)
library(pcadapt)
library(LEA)
library(vegan)
library(lfmm)
library(vcfR)
library(adegenet)
library(ade4)
library(tibble)
library(data.table)
library(topGO)
```

```{r}
col_pal <- c("#0072B2", "#56B4E9",  "#F0E442", "#E69F00")

```
# Mortality Data
```{r}
df.mortality <- read.csv("data/oyster_survival_data.csv", stringsAsFactors = FALSE)
# Keep only the columns needed for survival analysis and clean up
df.mortality <- df.mortality %>%
  select(Treatment, Time, Status) %>%
  filter(!is.na(Time), !is.na(Status)) %>%
  mutate(
    Treatment = factor(
      Treatment,
      levels = c("CONCON", "STRCON", "CONROD", "STRROD")
    ),
    # Status: 1 = event (death), 0 = censored
    Status = as.integer(Status)
  )
df.mortality$Treatment <- factor(df.mortality$Treatment, levels=c("CONCON", "STRCON", "CONROD", "STRROD"))
```

```{r}
km_fit <- survfit2(Surv(Time, Status) ~ Treatment, data = df.mortality)

km.plot <- km_fit %>%
  ggsurvfit(linewidth = 0.9) +
  add_confidence_interval() +
  add_pvalue(
    location   = "annotation",
    x          = 1,
    y          = 0.08,
    size       = 3.5,
    prepend_p  = TRUE
  ) +
  scale_color_manual(
    values = col_pal,
    labels = c("CONCON", "STRCON", "CONROD", "STRROD")
  ) +
  scale_fill_manual(
    values = col_pal,
    labels = c("CONCON", "STRCON", "CONROD", "STRROD")
  ) +
  scale_x_continuous(
    breaks = c(0, 5, 10, 15, 20, 22),
    limits = c(0, 24)
  ) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0, 1)
  ) +
  labs(
    title    = "Kaplan-Meier Survival Curves by Treatment",
    #subtitle = "Oyster survival across four treatment groups",
    x        = "Days",
    y        = "Cumulative Survival Probability",
    color    = "Treatment",
    fill     = "Treatment"
  ) +
  theme_ggsurvfit_default() +
  theme_classic(base_size = 13)+
  theme(
    plot.title      = element_text(face = "bold", size = 14),
    plot.subtitle   = element_text(size = 11, colour = "grey40"),
    legend.position = "right",
    legend.title    = element_text(face = "bold"),
    axis.title      = element_text(size = 11),
    axis.text       = element_text(size = 10)
  )

km.plot
```
```{r}
df.F.mortality <- df.mortality %>%
  group_by(Treatment) %>%
  summarise(
    Total     = n(),
    Deaths    = sum(Status == 1),
    Mortality = Deaths / Total * 100,      # as percentage
    SE        = sqrt((Mortality/100 * (1 - Mortality/100)) / Total) * 100
  ) %>%
  ungroup()


df.F.mortality$Treatment <- factor(
  df.F.mortality$Treatment,
  levels = c("CONCON", "STRCON", "CONROD", "STRROD")
)

finalmort.plot <- ggplot(df.F.mortality, aes(x = Treatment, y = Mortality, fill = Treatment)) +
  geom_col(width = 0.6, colour = "black", linewidth = 0.4, alpha=0.9) +
  geom_errorbar(aes(ymin = Mortality - SE, ymax = Mortality + SE),width    = 0.2,linewidth = 0.7) +
  scale_fill_manual(values = col_pal) +
  scale_y_continuous(
    limits = c(0, 80),
    breaks = seq(0, 80, 10),
    labels = function(x) paste0(x, "%"),
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(
    title    = "Final Mortality by Treatment Group",
    #subtitle = "Error bars represent ± 1 SE",
    x        = "Treatment",
    y        = "Mortality (%)"
  ) +
  theme_classic(base_size = 13) +
  theme(
    plot.title      = element_text(face = "bold", size = 14),
    plot.subtitle   = element_text(size = 11, colour = "grey40"),
    axis.title      = element_text(size = 12),
    axis.text       = element_text(size = 11),
    legend.position = "none"
  )


finalmort.plot
```


# Bioinformatics
## SNP calling
```{bash EVAL=FALSE}
mamba env create -f ROD_CADO.yml
mamba activate ROD_CADO

cd ./analysis

../scripts/dDocent_ngs dDocent.config
```
## SNP filtering
```{bash EVAL=FALSE}
mamba activate ROD_CADO

cd ./analysis/raw.vcf
../../scripts/PVCF_filter.sh TRS 36
cd filtered
vcftools --gzvcf SNP.TRS.TRSdp5g75.nDNA.g1.maf01.FIL.vcf.gz --minDP 10 --recode --recode-INFO-all --stdout | vcftools --vcf - --max-missing 1.0 --maf 0.01 --recode --recode-INFO-all --stdout | bcftools view --threads 24 -O z -o SNP.TRSdp10g1.FIL.vcf.gz
bcftools --threads 40 -m2 -M2 -O z -o SNP.TRSdp10g1max2alleles.FIL.vcf.gz SNP.TRSdp10g1.FIL.vcf.gz
plink2 --allow-extra-chr --make-bed --out TRSdp10g1.FIL --vcf SNP.TRSdp10g1max2alleles.FIL.vcf.gz
```

# Diversity Metrics

## Nucleotide diversity
Original analysis by Jacob Green, edits by JBP

### Calculate from vcf files
```{bash, EVAL =FALSE}
source activate ROD_CADO
mkdir -p ./analysis/diversity_statistics/nd
cd ./analysis/diversity_statistics/nd

cp ../../../treat_popmap .

awk '$2 == "CON-CON" {print $1}' treat_popmap > con-con.txt 
awk '$2 == "STR-CON" {print $1}' treat_popmap > str-con.txt 
awk '$2 == "CON-ROD" {print $1}' treat_popmap > con-rod.txt 
awk '$2 == "STR-ROD" {print $1}' treat_popmap > str-rod.txt

```


```{bash eval=FALSE}
source activate ROD_CADO
cd ./analysis/diversity_statistics/nd


bcftools view --threads 20 ../../raw.vcf/filtered/SNP.TRSdp10g1.FIL.vcf.gz -S con-con.txt | vcftools --vcf - --window-pi 10000 --out con-con

bcftools view --threads 20 ../../raw.vcf/filtered/SNP.TRSdp10g1.FIL.vcf.gz -S str-con.txt | vcftools --vcf - --window-pi 10000 --out str-con

bcftools view --threads 20 ../../raw.vcf/filtered/SNP.TRSdp10g1.FIL.vcf.gz -S con-rod.txt | vcftools --vcf - --window-pi 10000 --out con-rod

bcftools view --threads 20 ../../raw.vcf/filtered/SNP.TRSdp10g1.FIL.vcf.gz -S str-rod.txt | vcftools --vcf - --window-pi 10000 --out str-rod

```

### Load into R
```{r, message=FALSE, warning=FALSE}
pi.concon.dataframe<-read.table("analysis/diversity_statistics/nd/con-con.windowed.pi", sep="\t", header=T)
pi.conrod.dataframe<-read.table("analysis/diversity_statistics/nd/con-rod.windowed.pi", sep="\t", header=T)
pi.strcon.dataframe<-read.table("analysis/diversity_statistics/nd/str-con.windowed.pi", sep="\t", header=T)
pi.strrod.dataframe<-read.table("analysis/diversity_statistics/nd/str-rod.windowed.pi", sep="\t", header=T)

pi.concon.dataframe$Treatment <- "CONCON"
pi.conrod.dataframe$Treatment <- "CONROD"
pi.strcon.dataframe$Treatment <- "STRCON"
pi.strrod.dataframe$Treatment <- "STRROD"

pi.df <- rbind(pi.concon.dataframe,pi.conrod.dataframe,pi.strcon.dataframe, pi.strrod.dataframe)
pi.df$Treatment <- factor(pi.df$Treatment, levels=c("CONCON", "STRCON", "CONROD", "STRROD"))
```

```{r}
pi.plot <- ggplot(pi.df, aes(x=PI,y = Treatment))+
  geom_violin(aes(color=Treatment,fill= Treatment), alpha=0.9) +
  #stat_summary(fun=mean, geom="point", shape=23, size=4)+
  #geom_dotplot(binaxis='y', stackdir='center', dotsize=0.005)+
  #geom_boxplot(width=0.25,outlier.shape = 23, outlier.color = "black", fill = NA)+
  geom_boxplot(width=0.25,outlier.shape = NA, outlier.color = NA, fill = NA)+
  stat_summary(fun=mean, geom="point", shape=23, size=2)+
  scale_fill_manual(values=col_pal)+
  scale_color_manual(values=col_pal)+
  scale_y_discrete(limits = rev(levels(pi.df$Treatment)))+
  #ylab("Population/Line")+
  labs(x=expression(pi)) +
  ggtitle("Nucleotide Diversity by Treatment")+
  theme_classic(base_size = 13) +
  theme(axis.title.y = element_blank(),plot.title = element_text(face = "bold"),legend.title   = element_text(face = "bold")) 

pi.plot
```
```{r}
group_by(pi.df, Treatment) %>%
  dplyr::summarise(
    count = n(),
    mean = mean(PI, na.rm = TRUE),
    sd = sd(PI, na.rm = TRUE),
    se=std.error(PI, na.rm = TRUE) )
```


```{r}

aov.pi <- aov(PI ~ Treatment, pi.df)
summary(aov.pi)

posthoc.pi <- TukeyHSD(aov.pi)
posthoc.pi


```

## Heterozygosity
Analysis by Coline Caillon with edits by Jon Puritz

```{bash eval=FALSE}
source activate ROD_CADO
mkdir -p ./analysis/diversity_statistics/het
cd ./analysis/diversity_statistics/het

popStats -y GT --file <(bcftools view --threads 40 ../../raw.vcf/filtered/SNP.TRSdp10g1.FIL.vcf.gz) --target 0,1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19 > CONCON.het &

popStats -y GT --file <(bcftools view --threads 40 ../../raw.vcf/filtered/SNP.TRSdp10g1.FIL.vcf.gz) --target 20,21,22,23,24,25,26,27,28,29,30,31,32,33,34,35,36,37,38,39 > STRCON.het &

popStats -y GT --file <(bcftools view --threads 40 ../../raw.vcf/filtered/SNP.TRSdp10g1.FIL.vcf.gz) --target 40,41,42,43,44,45,46,47,48,49,50,51,52,53,54,55,56,57,58,59 > CONROD.het &

popStats -y GT --file <(bcftools view --threads 40 ../../raw.vcf/filtered/SNP.TRSdp10g1.FIL.vcf.gz) --target 60,61,62,63,64,65,66,67,68,69,70,71,72,73,74,75,76,77,78,79 > STRROD.het

```

```{bash eval=FALSE}
source activate ROD_CADO
cd ./analysis/diversity_statistics/het

# CON-CON
cat <(echo -e "CHROM\tBIN_START\tBIN_END\tN_SITES\tN_SNPs\tHET\tTREAT") <(mawk '{if (NR>1) print $1 "\t" $2-1 "\t" $2 "\t" $4 "\t" $5 "\tCONCON"}' CONCON.het) > CONCON-treat.het

# STR-CON
cat <(echo -e "CHROM\tBIN_START\tBIN_END\tN_SITES\tN_SNPs\tHET\tTREAT") <(mawk '{if (NR>1) print $1 "\t" $2-1 "\t" $2 "\t" $4 "\t" $5 "\tSTRCON"}' STRCON.het) > STRCON-treat.het

# CON-ROD
cat <(echo -e "CHROM\tBIN_START\tBIN_END\tN_SITES\tN_SNPs\tHET\tTREAT") <(mawk '{if (NR>1) print $1 "\t" $2-1 "\t" $2 "\t" $4 "\t" $5 "\tCONROD"}' CONROD.het) > CONROD-treat.het

# STR-ROD
cat <(echo -e "CHROM\tBIN_START\tBIN_END\tN_SITES\tN_SNPs\tHET\tTREAT") <(mawk '{if (NR>1) print $1 "\t" $2-1 "\t" $2 "\t" $4 "\t" $5 "\tSTRROD"}' STRROD.het) > STRROD-treat.het

```

```{bash}
source activate ROD_CADO
cd ./analysis/diversity_statistics/het
bedtools makewindows -g ../../../other_files/genome.file -w 10000 | mawk '!/NC_007175.2/' > 10kb.bed

bedtools intersect -b CONCON-treat.het -a 10kb.bed -wa -wb | bedtools merge -d -1 -c 7,8 -o mean > CONCON.het.10kb.bed
bedtools intersect -b CONROD-treat.het -a 10kb.bed -wa -wb | bedtools merge -d -1 -c 7,8 -o mean > CONROD.het.10kb.bed
bedtools intersect -b STRCON-treat.het -a 10kb.bed -wa -wb | bedtools merge -d -1 -c 7,8 -o mean > STRCON.het.10kb.bed
bedtools intersect -b STRROD-treat.het -a 10kb.bed -wa -wb | bedtools merge -d -1 -c 7,8 -o mean > STRROD.het.10kb.bed

cat <(echo -e "CHROM\tBIN_START\tBIN_END\tEXP_HET\tOBS_HET\tTREATMENT") <(mawk '{if (NR>1)print $0 "\tCONCON"}' CONCON.het.10kb.bed) <(mawk '{if (NR>1)print $0 "\tCONROD"}' CONROD.het.10kb.bed) <(mawk '{if (NR>1)print $0 "\tSTRCON"}' STRCON.het.10kb.bed) <(mawk '{if (NR>1)print $0 "\tSTRROD"}' STRROD.het.10kb.bed)  > treat.10kb.txt

```

### Load into R
```{r, message=FALSE, warning=FALSE}
het.df <- read.table("analysis/diversity_statistics/het/treat.10kb.txt", sep="\t", header=T)
het.df$Treatment <- factor(het.df$TREATMENT, levels=c("CONCON", "STRCON", "CONROD", "STRROD"))
```

```{r}
het.plot <- ggplot(het.df, aes(x=OBS_HET,y = Treatment))+
  geom_violin(aes(color=Treatment,fill= Treatment), alpha =0.9) +
  #geom_boxplot(width=0.25,outlier.shape = 23, outlier.color = "black", fill = NA)+
  geom_boxplot(width=0.25,outlier.shape = NA, outlier.color = NA, fill = NA)+
  stat_summary(fun=mean, geom="point", shape=23, size=2)+
  scale_fill_manual(values=col_pal)+
  scale_color_manual(values=col_pal)+
  scale_y_discrete(limits = rev(levels(het.df$Treatment)))+
  labs(x="Observed Heterozygosity") +
  ggtitle("Observed Heterozygosity by Treatment")+
  theme_classic(base_size = 13) +
    theme(axis.title.y = element_blank(),plot.title = element_text(face = "bold"),legend.title   = element_text(face = "bold")) 
het.plot
```
```{r}
group_by(het.df, Treatment) %>%
  dplyr::summarise(
    count = n(),
    mean = mean(OBS_HET, na.rm = TRUE),
    sd = sd(OBS_HET, na.rm = TRUE),
    se=std.error(OBS_HET) )
```


```{r}

aov.het <- aov(OBS_HET ~ Treatment, het.df)
summary(aov.het)

posthoc.het <- TukeyHSD(aov.het)
posthoc.het


```

## Tajima's D
Code by Megan Guidry with edits by Jon Puritz
```{bash, eval = FALSE}
mamba create -n vkrc python=3.8 numpy scipy
source activate vkrc
pip install VCF-kit
mamba install bcftools
```


```{bash eval=FALSE}
source activate vkrc
mkdir -p ./analysis/diversity_statistics/td
cd ./analysis/diversity_statistics/td

bcftools view --threads 20 ../../raw.vcf/filtered/SNP.TRSdp10g1.FIL.vcf.gz -S ../nd/con-con.txt | vk tajima 10000 10000 - > con-con.tajima &

bcftools view --threads 20 ../../raw.vcf/filtered/SNP.TRSdp10g1.FIL.vcf.gz -S ../nd/str-con.txt | vk tajima 10000 10000 - > str-con.tajima &

bcftools view --threads 20 ../../raw.vcf/filtered/SNP.TRSdp10g1.FIL.vcf.gz -S ../nd/con-rod.txt | vk tajima 10000 10000 - > con-rod.tajima &

bcftools view --threads 20 ../../raw.vcf/filtered/SNP.TRSdp10g1.FIL.vcf.gz -S ../nd/str-rod.txt | vk tajima 10000 10000 - > str-rod.tajima
```


```{bash}
source activate ROD_CADO
cd ./analysis/diversity_statistics/td
cat <(echo -e "CHROM\tBIN_START\tBIN_END\tN_SITES\tN_SNPs\tTAJD\tTREATMENT") <(mawk '{if (NR>1)print $0 "\tCONCON"}' con-con.tajima) <(mawk '{if (NR>1)print $0 "\tSTRCON"}' str-con.tajima) <(mawk '{if (NR>1)print $0 "\tCONROD"}' con-rod.tajima) <(mawk '{if (NR>1)print $0 "\tSTRROD"}' str-rod.tajima) > treatment.tajima
```

### Load into R
```{r, message=FALSE, warning=FALSE}
tajd.df <- read.table("analysis/diversity_statistics/td/treatment.tajima", sep="\t", header=T)
tajd.df$Treatment <- factor(tajd.df$TREATMENT, levels=c("CONCON", "STRCON", "CONROD", "STRROD"))
```

```{r}
tajd.plot <- ggplot(tajd.df, aes(x=TAJD,y = Treatment))+
  geom_violin(aes(color=Treatment,fill= Treatment)) +
  #stat_summary(fun=mean, geom="point", shape=23, size=4)+
  #geom_dotplot(binaxis='y', stackdir='center', dotsize=0.005)+
  #geom_boxplot(width=0.25,outlier.shape = 23, outlier.color = "black", fill = NA)+
  geom_boxplot(width=0.25,outlier.shape = NA, outlier.color = NA, fill = NA)+
  stat_summary(fun=mean, geom="point", shape=23, size=2)+
  scale_fill_manual(values=col_pal)+
  scale_color_manual(values=col_pal)+
  scale_y_discrete(limits = rev(levels(tajd.df$Treatment)))+
  #ylab("Population/Line")+
  labs(x = expression("Tajima's "*italic(D))) +
  ggtitle(expression("Tajima's "*italic(D)*" by Treatment")) +
  theme_classic(base_size = 13) +
  theme(axis.title.y = element_blank(), text = element_text(family = "Liberation Sans"))

tajd.plot


```

```{r}
group_by(tajd.df, Treatment) %>%
  dplyr::summarise(
    count = n(),
    mean = mean(TAJD, na.rm = TRUE),
    sd = sd(TAJD, na.rm = TRUE),
    se=std.error(TAJD) )
```

```{r}

aov.tajd <- aov(TAJD ~ Treatment, tajd.df)
summary(aov.tajd)

posthoc.tajd <- TukeyHSD(aov.tajd)
posthoc.tajd

```

# PCA
Original Code by Jacob Green, edits by Jon Puritz
```{bash, eval = FALSE}
cd analysis/raw.vcf/filtered/
plink2 --bfile TRSdp10g1.FIL --make-bed --allow-extra-chr --out TRSdp10g1.FIL.ALL --maf 0.05

plink2 --bfile TRSdp10g1.FIL --make-bed --allow-extra-chr --keep <(cut -f1 ../../popmap | grep -v CH | grep -v R[1,2,3] ) --out TRSdp10g1.FIL.CMB --maf 0.05

plink2 --bfile TRSdp10g1.FIL --make-bed --allow-extra-chr --keep <(cut -f1 ../../popmap | grep -v CH | grep -v RH ) --out TRSdp10g1.FIL.ROD --maf 0.05

plink2 --bfile TRSdp10g1.FIL --make-bed --allow-extra-chr --keep <(cut -f1 ../../popmap | grep -v R ) --out TRSdp10g1.FIL.STR --maf 0.05
```

```{r}
#Set path to file
path_to_file <- "analysis/raw.vcf/filtered/TRSdp10g1.FIL.bed"
#Load file to filename variable
filename <- read.pcadapt(path_to_file, type = "bed")

resld <- pcadapt(filename, K = 10, LD.clumping = list(size = 500, thr = 0.2))
plot(resld, option = "screeplot")
```

```{r}
poplist.names <- c(rep("CONCON",20),rep("STRCON",20),rep("CONROD",20),rep("STRROD",20))
poplist.names
```

```{r}
df.pca <- as.data.frame(resld$scores)
colnames(df.pca) <- paste0("PC", 1:ncol(df.pca))

df.pca$Treatment <- poplist.names
df.pca$Treatment <- factor(df.pca$Treatment, levels=c("CONCON", "STRCON", "CONROD", "STRROD"))

variance_explained <- res$singular.values^2 / sum(res$singular.values^2) * 100
cat("Variance explained by each PC:\n")
print(round(variance_explained, 2))

pca.plot <- ggplot(df.pca, aes(x = PC1, y = PC2, colour = Treatment)) +
  geom_point(size = 2.5, alpha = 0.85) +
  labs(
    title = "PCA of Genetic Variation",
    x = paste0("PC1 (", round(variance_explained[1], 1), "%)"),
    y = paste0("PC2 (", round(variance_explained[2], 1), "%)")
  ) +
  theme_classic(base_size = 13) +
  scale_fill_manual(values=col_pal)+
  scale_color_manual(values=col_pal)+
  theme(
    plot.title     = element_text(face = "bold"),
    legend.title   = element_text(face = "bold"),
    legend.position = "right"
  )

pca.plot
```

# Figure 1
```{bash}
mkdir -p figures
```

```{r}
png(filename="figures/Figure1.png", type="cairo",units="px", width=5600, height=3000, res=300, bg="transparent")

gen_div.plot <- (pi.plot / het.plot)

(finalmort.plot + pca.plot)/(km.plot + gen_div.plot) + plot_layout(guides="collect") + plot_annotation(tag_levels = 'A')
dev.off()
```

# Outlier Detection
Original analysis by Amy Zyck, edits by JBP

## Setup

### Prepare genotype file for LFMM2
```{bash eval=FALSE}
source activate ROD_CADO
mkdir -p ./analysis/outlier_detection
cd ./analysis/outlier_detection
pwd

plink2 --vcf ../raw.vcf/filtered/SNP.TRSdp10g1.FIL.vcf.gz --allow-extra-chr --min-alleles 2 --max-alleles 2 --export A --out snp_rod
cut -f7- snp_rod.raw | tail -n +2 | sed 's/NA/9/g' > snp_rod.lfmm
rm snp_rod.raw
```

### Prepare environmental strata
```{bash eval=FALSE}
cd ./analysis/outlier_detection
cp -f ../../other_files/strata_ROD_tab .
```

### Create SNP index file
```{bash eval=FALSE}
source activate ROD_CADO
cd ./analysis/outlier_detection

bcftools view -m 2 -M 2 ../raw.vcf/filtered/SNP.TRSdp10g1.FIL.vcf.gz | bcftools query -f '%CHROM\t%POS\n'  > totalloci
NUM=(`cat totalloci | wc -l`)
paste <(seq 1 $NUM) totalloci > loci.plus.index
```

## Run LFMM2


### Load genotype and environmental data
```{r}
clim.env_rod <- read.table("analysis/outlier_detection/strata_ROD_tab", header=TRUE)
gen_rod <- fread("analysis/outlier_detection/snp_rod.lfmm", header = FALSE)
row.names(gen_rod) <- clim.env_rod$Individual

CADO <- clim.env_rod$CADO
DIS <- clim.env_rod$Disease
COMB_BIN <- clim.env_rod$Combined_Bin
```

### LFMM ridge - Stress (CADO)
```{r}
oys_rod_CADO.lfmm <- lfmm_ridge(Y=gen_rod, X=CADO, K=4)
oys_rod_CADO.pv <- lfmm_test(Y=gen_rod, X=CADO, lfmm=oys_rod_CADO.lfmm, calibrate="gif")
oys_rod_CADO.pv$gif
```

```{r}
zscore_rod_CADO <- oys_rod_CADO.pv$score[,1]
adj.pv_rod_CADO <- pchisq(zscore_rod_CADO^2/0.98, df=1, lower = FALSE)
oys_rod_CADO.qv <- qvalue(adj.pv_rod_CADO)$qvalues
oys_rod_CADO.FDR.1 <- as.vector(which(oys_rod_CADO.qv < 0.1))
length(oys_rod_CADO.FDR.1)
```

```{r}
cat("", file = "analysis/outlier_detection/outliers_lfmm2_CADO_098.txt")
invisible(lapply(oys_rod_CADO.FDR.1, write, "analysis/outlier_detection/outliers_lfmm2_CADO_098.txt", append=TRUE))
```

### LFMM ridge - Disease
```{r}
oys_rod_DIS.lfmm <- lfmm_ridge(Y=gen_rod, X=DIS, K=4)
oys_rod_DIS.pv <- lfmm_test(Y=gen_rod, X=DIS, lfmm=oys_rod_DIS.lfmm, calibrate="gif")
oys_rod_DIS.pv$gif
```

```{r}
zscore_rod_DIS <- oys_rod_DIS.pv$score[,1]
adj.pv_rod_DIS <- pchisq(zscore_rod_DIS^2/0.98, df=1, lower = FALSE)
oys_rod_DIS.qv <- qvalue(adj.pv_rod_DIS)$qvalues
oys_rod_DIS.FDR.1 <- as.vector(which(oys_rod_DIS.qv < 0.1))
length(oys_rod_DIS.FDR.1)
```

```{r}
cat("", file = "analysis/outlier_detection/outliers_lfmm2_DIS_098.txt")
invisible(lapply(oys_rod_DIS.FDR.1, write, "analysis/outlier_detection/outliers_lfmm2_DIS_098.txt", append=TRUE))
```

### LFMM ridge - Combined treatment (binary)
```{r}
oys_rod_CB.lfmm <- lfmm_ridge(Y=gen_rod, X=COMB_BIN, K=4)
oys_rod_CB.pv <- lfmm_test(Y=gen_rod, X=COMB_BIN, lfmm=oys_rod_CB.lfmm, calibrate="gif")
oys_rod_CB.pv$gif
```

```{r}
oys_rod_CB.qv <- qvalue(oys_rod_CB.pv$calibrated.pvalue)$qvalues
oys_rod_CB.FDR.1 <- which(oys_rod_CB.qv < 0.1)
length(oys_rod_CB.FDR.1)
```

```{r}
cat("", file = "analysis/outlier_detection/outliers_lfmm2_CB.txt")
invisible(lapply(oys_rod_CB.FDR.1, write, "analysis/outlier_detection/outliers_lfmm2_CB.txt", append=TRUE))
```

## Combine outliers and create BED file

### Map SNP indices to genomic positions
```{bash eval=FALSE}
source activate ROD_CADO
mkdir -p analysis/outlier_detection/results
cd ./analysis/outlier_detection

cat outliers_lfmm2_CADO_098.txt | parallel "grep -w ^{} loci.plus.index" | cut -f2,3 > ./results/outliers_lfmm2_CADO_098.loci.txt
cat outliers_lfmm2_DIS_098.txt | parallel "grep -w ^{} loci.plus.index" | cut -f2,3 > ./results/outliers_lfmm2_DIS_098.loci.txt
cat outliers_lfmm2_CB.txt | parallel "grep -w ^{} loci.plus.index" | cut -f2,3 > ./results/outliers_lfmm2_CB.loci.txt

cat ./results/outliers_lfmm2_CB.loci.txt ./results/outliers_lfmm2_CADO_098.loci.txt ./results/outliers_lfmm2_DIS_098.loci.txt > ./results/all_lfmm2_outliers
awk '{print $1 "\t" $2-1 "\t" $2}' ./results/all_lfmm2_outliers | bedtools sort -i - > ./results/outliers_lfmm2.bed
```

### Create outlier VCF for validation
```{bash eval=FALSE}
source activate ROD_CADO
cd ./analysis/outlier_detection

bcftools index --threads 20 ../raw.vcf/filtered/SNP.TRSdp10g1.FIL.vcf.gz

bcftools view --threads 20 -R ./results/outliers_lfmm2_CB.loci.txt ../raw.vcf/filtered/SNP.TRSdp10g1.FIL.vcf.gz -O b -o outliers_lfmm2_CB.recode.vcf
```

## Validate outliers with PCA and PERMANOVA

```{r}
strata_ROD <- read.table("analysis/outlier_detection/strata_ROD_tab", header=TRUE)
pop <- poplist.names
Combined <- c(rep("No", 20),rep("No", 20),rep("No", 20),rep("Yes", 20))
```

```{r}
vcf_out_lfmm_CB <- read.vcfR("analysis/outlier_detection/outliers_lfmm2_CB.recode.vcf")
rad_out_lfmm_CB.filt <- vcfR2genind(vcf_out_lfmm_CB, strata = strata_ROD, pop = pop)
```

```{r}
x_out_lfmm_CB <- scaleGen(rad_out_lfmm_CB.filt, NA.method = "mean")
out_lfmm_CB_pca <- dudi.pca(x_out_lfmm_CB, center = FALSE, scale = FALSE, scannf = FALSE, nf = 4)
pc_scores_CB <- out_lfmm_CB_pca$li
```

```{r}
permanova_result_CB <- adonis2(pc_scores_CB ~ pop, method = "euclidean")
print(permanova_result_CB)
```

```{r}
permanova_result_CB_combined <- adonis2(pc_scores_CB ~ Combined, method = "euclidean")
print(permanova_result_CB_combined)
```

```{r}

pca_data_CB <- data.frame(
  PC1 = pc_scores_CB[,1],
  PC2 = pc_scores_CB[,2],
  Treatment = factor(poplist.names, levels = c("CONCON", "STRCON", "CONROD", "STRROD"))
)

p_value_CB <- permanova_result_CB$`Pr(>F)`[1]
eigenvalues_CB <- eigen(cov(pca_data_CB[, c("PC1", "PC2")]))$values
pve_CB <- eigenvalues_CB / sum(eigenvalues_CB) * 100

outlier.pca.plot <- ggplot(pca_data_CB, aes(x = PC1, y = PC2, color = Treatment)) +
  geom_point(size = 2.5, alpha = 0.8) +
  stat_ellipse() +
  scale_color_manual(values = col_pal) +
  labs(
    title = "PCA of Combined Treatment Outlier SNPs from LFMM",
    x = paste0("PC1 (", round(pve_CB[1], 1), "%)"),
    y = paste0("PC2 (", round(pve_CB[2], 1), "%)")
  ) +
  annotate("text", x = -Inf, y = -Inf,
           label = paste("PERMANOVA p =", format(p_value_CB, digits = 2)),
           hjust = -0.1, vjust = -1, size = 3.5) +
  theme_classic(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    legend.position = "right"
  )

outlier.pca.plot
```

# FST and OutFLANK
```{bash, eval = FALSE}
source activate ROD_CADO
mkdir -p ./analysis/FST
cd ./analysis/FST

plink2 --bfile ../raw.vcf/filtered/TRSdp10g1.FIL --make-bed --allow-extra-chr --out TRSdp10g1.FIL.ALL --maf 0.05

```








```{r}
# Read in bedfile
rds <- snp_readBed("./analysis/raw.vcf/filtered/TRSdp10g1.FIL.ALL.bed", backingfile = tempfile())
```

```{r}
full_bed <- snp_attach(rds)
G <- full_bed$genotypes
CHR <- full_bed$map$chromosome
POS <- full_bed$map$physical.pos
NCORES <- nb_cores()
```

```{r}
svd <- snp_autoSVD(G, as.integer(factor(CHR)), POS, ncores = NCORES,min.mac =8,size=10)
```

```{r}
write(paste(CHR[attr(svd, "subset")],
            POS[attr(svd, "subset")], sep='\t'),
      "./analysis/FST/ThinnedSNPs.maskedPCs.txt")
```

```{bash, eval = FALSE}
source activate ROD_CADO
cd ./analysis/FST
rm -f random.50k.snps

cut -f1 ThinnedSNPs.maskedPCs.txt| uniq | parallel "grep {} ThinnedSNPs.maskedPCs.txt | shuf  2>/dev/null | shuf 2>/dev/null | shuf 2>/dev/null| head -5000" >> random.50k.snps

```

```{bash, eval = FALSE}
source activate ROD_CADO
cd ./analysis/FST

plink2 --bfile ../raw.vcf/filtered/TRSdp10g1.FIL --make-bed --allow-extra-chr --extract bed0 <(mawk '{print $1"\t"$2-1"\t"$2}' random.50k.snps ) --out random50k

mawk 'NR==1{print "IND\tPOP"};NR >0 {print $0}' ../popmap > popmap.txt
```

```{r}
rds.random <- snp_readBed("random.50k.snps", backingfile = tempfile())

random <- snp_attach(rds.random)
G.r <- random$genotypes
CHR.r <- random$map$chromosome
POS.r <- random$map$physical.pos

```

```{r}

popmap <- read.table("./analysis/FST/popmap.txt", header=TRUE)
popmap$TREAT <- c(rep("CON",20),rep("STR",20),rep("ROD",20),rep("STRROD",20))


my_fst <- MakeDiploidFSTMat(G[], locusNames = paste0(CHR,"_", POS), popNames = popmap$TREAT)

my_fst.r <- MakeDiploidFSTMat(G.r[], locusNames = paste0(CHR.r,"_", POS.r), popNames = popmap$TREAT)




```




```{r}
save(my_fst, file = "./analysis/FST/FST_data.RData")
```

```{r}
load("./analysis/FST/FST_data.RData")

```


```{r}
out_trim <- OutFLANK(my_fst[attr(svd, which="subset"),], NumberOfSamples=4, qthreshold = 0.1, Hmin = 0.1)

OutFLANKResultsPlotter(out_trim, withOutliers = TRUE,
                       NoCorr = TRUE, Hmin = 0.1, binwidth = 0.001, Zoom =
                         FALSE, RightZoomFraction = 0.05, titletext = NULL)

hist(out_trim$results$pvaluesRightTail)

qthres <- 0.1

P1 <- pOutlierFinderChiSqNoCorr(my_fst, Fstbar = out_trim$FSTNoCorrbar, 
                                   dfInferred = out_trim$dfInferred, qthreshold = qthres, Hmin=0.1)

P1 <- P1[order(as.numeric(rownames(P1))),,drop=FALSE]                  
hist(P1$pvaluesRightTail)
```


```{r}
plot.df<- setNames(data.frame(matrix(ncol = 8, nrow = length(P1$LocusName))), c("CHR", "CHRR", "BP", "SNP", "P", "F", "Q", "Outlier"))
plot.df$CHR <-as.integer(as.factor(CHR))
plot.df$CHRR <- CHR
plot.df$SNP <-P1$LocusName
plot.df$BP <- POS
plot.df$P <- P1$pvalues
plot.df$Q <- P1$qvalues
plot.df$F <- P1$FST
plot.df$Outlier <- P1$OutlierFlag
plot.df <- plot.df[which(plot.df$Outlier == TRUE | plot.df$Outlier == FALSE),]


```

```{r}
withr::with_options(c(scipen = 10), write(paste(plot.df$CHRR,plot.df$BP-1,plot.df$BP,plot.df$F,plot.df$Outlier, sep='\t'), "./analysis/FST/FST_Outflank.bed"))
```

```{bash}
source activate ROD_CADO
cd ./analysis/FST
bedtools intersect -b <(sort -k 1,1 -k2,2n FST_Outflank.bed) -a <(sort -k 1,1 -k2,2n ../diversity_statistics/het/10kb.bed) -wa -wb -sorted  |bedtools merge -d -1 -c 7,8 -o mean,distinct | mawk 'BEGIN {print "CHROM\tBIN_START\tBIN_END\tWEIGHTED_FST\tOUTLIER"} { if ($0 ~ /TRUE/) {print $1 "\t" $2 "\t" $3 "\t" $4 "\tTRUE"} else {print $0}}' > OF_fst.10kb.bed

cat <(echo -e "CHROM\tBIN_START\tBIN_END\tWEIGHTED_FST\tOUTLIER") OF_fst.10kb.bed <(grep TRUE FST_Outflank.bed) > OF_fst.10kb.txt



cat <(echo -e "CHROM\tBIN_START\tBIN_END\tWEIGHTED_FST\tOUTLIER") FST_Outflank.bed > outlier_fst.txt
```


```{r}
fst <- fread("./analysis/FST/outlier_fst.txt", header = TRUE)
SNP<-c(1:(nrow(fst)))
mydf<-data.frame(SNP,fst)
mydf$CHR <- mydf$CHROM
mydf %>% 
  mutate(CHR = str_replace(CHR, "NC_035780.1", "1")) %>% 
  mutate(CHR = str_replace(CHR, "NC_035781.1", "2")) %>% 
  mutate(CHR = str_replace(CHR, "NC_035782.1", "3")) %>% 
  mutate(CHR = str_replace(CHR, "NC_035783.1", "4")) %>% 
  mutate(CHR = str_replace(CHR, "NC_035784.1", "5")) %>% 
  mutate(CHR = str_replace(CHR, "NC_035785.1", "6")) %>% 
  mutate(CHR = str_replace(CHR, "NC_035786.1", "7")) %>% 
  mutate(CHR = str_replace(CHR, "NC_035787.1", "8")) %>%
  mutate(CHR = str_replace(CHR, "NC_035788.1", "9")) %>% 
  mutate(CHR = str_replace(CHR, "NC_035789.1", "10"))  -> mydf
mydf$CHR <- as.numeric(mydf$CHR)  

# Step 1: Compute chromosome-level metrics
chr_offsets <- mydf %>%
  group_by(CHR) %>%
  summarize(
    minbp = min(BIN_START, na.rm = TRUE),
    maxbp = max(BIN_END, na.rm = TRUE),
    nrows = n()
  ) %>%
  mutate(
    offset = cumsum(nrows) - nrows,
    bp_range = ifelse(maxbp == minbp, 1, maxbp - minbp)
  )

# Step 2: Join offset info into the original dataframe and compute BPcum
don <- mydf %>%
  left_join(chr_offsets, by = "CHR") %>%
  mutate(
    scaled_bp = (as.numeric(BIN_START) - as.numeric(minbp)) * as.numeric(nrows) / as.numeric(bp_range),
    BPcum = scaled_bp + as.numeric(offset)
  )

# Step 3: Axis label positions
axisdf <- don %>%
  group_by(CHR) %>%
  summarize(center = mean(BPcum, na.rm = TRUE))




m2<- ggplot(don, aes(x=BPcum, y=WEIGHTED_FST)) +
  geom_point( aes(color=as.factor(CHR)), alpha=0.25, size=0.5) +
  geom_point(data=subset(don, OUTLIER==TRUE),aes(color=as.factor(CHR)),shape = 17,alpha=1,size=1.75) +
  scale_color_manual(values = rep(c("#7d5577", "#52374e"), 5 )) +
  scale_x_continuous( label = axisdf$CHR, breaks= axisdf$center ,expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0),limits=c(min(don$WEIGHTED_FST-0.001,na.rm=TRUE),0.5))+      
  theme(axis.title.y = element_markdown())+ 
  guides(colour = "none",alpha="none", fill="none") +
  labs(x = "Chromosome", y = "*F<sub>ST</sub>*") +
  theme_classic() +
  theme(axis.title.y = element_markdown()) 
    

png(filename="figures/Figure.X.FST.man.png", type="cairo",units="px", width=5600, height=3000, res=300, bg="transparent")
m2
null <- dev.off()

```

# Outlier Overlap and Annotation
Original analysis by Amy Zyck, edits by JBP

## Extract OutFLANK outliers to BED format
```{bash eval=FALSE}
source activate ROD_CADO
cd ./analysis/FST

awk '$5=="TRUE"{print $1"\t"$2"\t"$3}' FST_Outflank.bed > Outflank_outliers.bed
```

## Compare overlapping outlier SNPs between LFMM2 and OutFLANK

```{bash eval=FALSE}
source activate ROD_CADO
cd ./analysis

# Sort each bed file
sort ./FST/Outflank_outliers.bed > ./FST/outliers_outflank_sorted.txt
sort ./outlier_detection/results/outliers_lfmm2.bed > ./outlier_detection/results/outliers_lfmm2_sorted.txt
```

```{bash eval=FALSE}
cd ./analysis

# Shared outlier SNPs between OutFLANK and LFMM2
comm -12 ./FST/outliers_outflank_sorted.txt ./outlier_detection/results/outliers_lfmm2_sorted.txt | wc -l
comm -12 ./FST/outliers_outflank_sorted.txt ./outlier_detection/results/outliers_lfmm2_sorted.txt
```

## Validate all LFMM2 outliers with PCA and PERMANOVA

### Create VCF for all LFMM2 outliers
```{bash eval=FALSE}
source activate ROD_CADO
cd ./analysis/outlier_detection

bcftools view --threads 20 -R ./results/outliers_lfmm2.bed ../raw.vcf/filtered/SNP.TRSdp10g1.FIL.vcf.gz > outliers_lfmm2_all.recode.vcf

```

### Load and analyze all LFMM2 outliers
```{r}
vcf_out_lfmm_all <- read.vcfR("analysis/outlier_detection/outliers_lfmm2_all.recode.vcf")
rad_out_lfmm_all.filt <- vcfR2genind(vcf_out_lfmm_all, strata = strata_ROD, pop = pop)
```

```{r}
x_out_lfmm_all <- scaleGen(rad_out_lfmm_all.filt, NA.method = "mean")
out_lfmm_all_pca <- dudi.pca(x_out_lfmm_all, center = FALSE, scale = FALSE, scannf = FALSE, nf = 4)
pc_scores_lfmm_all <- out_lfmm_all_pca$li
```

```{r}
permanova_result_lfmm_all <- adonis2(pc_scores_lfmm_all ~ pop, method = "euclidean")
print(permanova_result_lfmm_all)
```

```{r}
permanova_result_lfmm_all_combined <- adonis2(pc_scores_lfmm_all ~ Combined, method = "euclidean")
print(permanova_result_lfmm_all_combined)
```

```{r}
pca_data_lfmm_all <- data.frame(
  PC1 = pc_scores_lfmm_all[,1],
  PC2 = pc_scores_lfmm_all[,2],
  Treatment = factor(poplist.names, levels = c("CONCON", "STRCON", "CONROD", "STRROD"))
)

p_value_lfmm_all <- permanova_result_lfmm_all$`Pr(>F)`[1]
eigenvalues_lfmm_all <- eigen(cov(pca_data_lfmm_all[, c("PC1", "PC2")]))$values
pve_lfmm_all <- eigenvalues_lfmm_all / sum(eigenvalues_lfmm_all) * 100

outlier.pca.lfmm.all.plot <- ggplot(pca_data_lfmm_all, aes(x = PC1, y = PC2, color = Treatment)) +
  geom_point(size = 2.5, alpha = 0.8) +
  stat_ellipse() +
  scale_color_manual(values = col_pal) +
  labs(
    title = "PCA of All LFMM2 Outlier SNPs",
    x = paste0("PC1 (", round(pve_lfmm_all[1], 1), "%)"),
    y = paste0("PC2 (", round(pve_lfmm_all[2], 1), "%)")
  ) +
  annotate("text", x = -Inf, y = -Inf,
           label = paste("PERMANOVA p =", format(p_value_lfmm_all, digits = 2)),
           hjust = -0.1, vjust = -1, size = 3.5) +
  theme_classic(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    legend.position = "right"
  )

outlier.pca.lfmm.all.plot
```

## Validate OutFLANK outliers with PCA and PERMANOVA

### Create VCF for OutFLANK outliers
```{bash eval=FALSE}
source activate ROD_CADO
cd ./analysis/FST

bcftools view --threads 20 -R Outflank_outliers.bed ../raw.vcf/filtered/SNP.TRSdp10g1.FIL.vcf.gz > outliers_Outflank.recode.vcf
```

### Load and analyze OutFLANK outliers
```{r}
vcf_out_outflank <- read.vcfR("analysis/FST/outliers_Outflank.recode.vcf")
rad_out_outflank.filt <- vcfR2genind(vcf_out_outflank, strata = strata_ROD, pop = pop)
```

```{r}
x_out_outflank <- scaleGen(rad_out_outflank.filt, NA.method = "mean")
out_outflank_pca <- dudi.pca(x_out_outflank, center = FALSE, scale = FALSE, scannf = FALSE, nf = 4)
pc_scores_outflank <- out_outflank_pca$li
```

```{r}
permanova_result_outflank <- adonis2(pc_scores_outflank ~ pop, method = "euclidean")
print(permanova_result_outflank)
```

```{r}
pca_data_outflank <- data.frame(
  PC1 = pc_scores_outflank[,1],
  PC2 = pc_scores_outflank[,2],
  Treatment = factor(poplist.names, levels = c("CONCON", "STRCON", "CONROD", "STRROD"))
)

p_value_outflank <- permanova_result_outflank$`Pr(>F)`[1]
eigenvalues_outflank <- eigen(cov(pca_data_outflank[, c("PC1", "PC2")]))$values
pve_outflank <- eigenvalues_outflank / sum(eigenvalues_outflank) * 100

outlier.pca.outflank.plot <- ggplot(pca_data_outflank, aes(x = PC1, y = PC2, color = Treatment)) +
  geom_point(size = 2.5, alpha = 0.8) +
  stat_ellipse() +
  scale_color_manual(values = col_pal) +
  labs(
    title = "PCA of OutFLANK Outlier SNPs",
    x = paste0("PC1 (", round(pve_outflank[1], 1), "%)"),
    y = paste0("PC2 (", round(pve_outflank[2], 1), "%)")
  ) +
  annotate("text", x = -Inf, y = -Inf,
           label = paste("PERMANOVA p =", format(p_value_outflank, digits = 2)),
           hjust = -0.1, vjust = -1, size = 3.5) +
  theme_classic(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    legend.position = "right"
  )

outlier.pca.outflank.plot
```

## PCA of combined LFMM2 and OutFLANK outliers

### Create combined outlier BED and VCF
```{bash eval=FALSE}
source activate ROD_CADO
cd ./analysis

# Merge LFMM2 and OutFLANK outlier BEDs, sort and deduplicate
cat ./outlier_detection/results/outliers_lfmm2.bed ./FST/Outflank_outliers.bed | bedtools sort -i - | bedtools merge -i - > ./outlier_detection/results/outliers_lfmm2_outflank_combined.bed

bcftools view --threads 20 -R ./outlier_detection/results/outliers_lfmm2_outflank_combined.bed ./raw.vcf/filtered/SNP.TRSdp10g1.FIL.vcf.gz -O v -o ./outlier_detection/outliers_combined.recode.vcf
```

### Load and analyze combined outliers
```{r}
vcf_out_combined <- read.vcfR("analysis/outlier_detection/outliers_combined.recode.vcf")
rad_out_combined.filt <- vcfR2genind(vcf_out_combined, strata = strata_ROD, pop = pop)
```

```{r}
x_out_combined <- scaleGen(rad_out_combined.filt, NA.method = "mean")
out_combined_pca <- dudi.pca(x_out_combined, center = FALSE, scale = FALSE, scannf = FALSE, nf = 4)
pc_scores_combined <- out_combined_pca$li
```

```{r}
permanova_result_combined <- adonis2(pc_scores_combined ~ pop, method = "euclidean")
print(permanova_result_combined)
```

```{r}
pca_data_combined <- data.frame(
  PC1 = pc_scores_combined[,1],
  PC2 = pc_scores_combined[,2],
  Treatment = factor(poplist.names, levels = c("CONCON", "STRCON", "CONROD", "STRROD"))
)

p_value_combined <- permanova_result_combined$`Pr(>F)`[1]
eigenvalues_combined <- eigen(cov(pca_data_combined[, c("PC1", "PC2")]))$values
pve_combined <- eigenvalues_combined / sum(eigenvalues_combined) * 100

outlier.pca.combined.plot <- ggplot(pca_data_combined, aes(x = PC1, y = PC2, color = Treatment)) +
  geom_point(size = 2.5, alpha = 0.8) +
  stat_ellipse() +
  scale_color_manual(values = col_pal) +
  labs(
    title = "PCA of Combined LFMM2 + OutFLANK Outlier SNPs",
    x = paste0("PC1 (", round(pve_combined[1], 1), "%)"),
    y = paste0("PC2 (", round(pve_combined[2], 1), "%)")
  ) +
  annotate("text", x = -Inf, y = -Inf,
           label = paste("PERMANOVA p =", format(p_value_combined, digits = 2)),
           hjust = -0.1, vjust = -1, size = 3.5) +
  theme_classic(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    legend.position = "right"
  )

outlier.pca.combined.plot
```

## PCA of overlapping LFMM2 and OutFLANK outlier loci

### Create overlapping outlier BED and VCF
```{bash eval=FALSE}
source activate ROD_CADO
cd ./analysis

# Extract loci shared between LFMM2 and OutFLANK
bedtools intersect -a ./outlier_detection/results/outliers_lfmm2.bed -b ./FST/Outflank_outliers.bed > ./outlier_detection/results/outliers_overlap_lfmm2_outflank.bed

wc -l ./outlier_detection/results/outliers_overlap_lfmm2_outflank.bed

bcftools view --threads 20 -R ./outlier_detection/results/outliers_overlap_lfmm2_outflank.bed ./raw.vcf/filtered/SNP.TRSdp10g1.FIL.vcf.gz -O v -o ./outlier_detection/outliers_overlap.recode.vcf
```

### Load and analyze overlapping outliers
```{r}
vcf_out_overlap <- read.vcfR("analysis/outlier_detection/outliers_overlap.recode.vcf")
rad_out_overlap.filt <- vcfR2genind(vcf_out_overlap, strata = strata_ROD, pop = pop)
```

```{r}
x_out_overlap <- scaleGen(rad_out_overlap.filt, NA.method = "mean")
out_overlap_pca <- dudi.pca(x_out_overlap, center = FALSE, scale = FALSE, scannf = FALSE, nf = 4)
pc_scores_overlap <- out_overlap_pca$li
```

```{r}
permanova_result_overlap <- adonis2(pc_scores_overlap ~ pop, method = "euclidean")
print(permanova_result_overlap)
```

```{r}
pca_data_overlap <- data.frame(
  PC1 = pc_scores_overlap[,1],
  PC2 = pc_scores_overlap[,2],
  Treatment = factor(poplist.names, levels = c("CONCON", "STRCON", "CONROD", "STRROD"))
)

p_value_overlap <- permanova_result_overlap$`Pr(>F)`[1]
eigenvalues_overlap <- eigen(cov(pca_data_overlap[, c("PC1", "PC2")]))$values
pve_overlap <- eigenvalues_overlap / sum(eigenvalues_overlap) * 100

outlier.pca.overlap.plot <- ggplot(pca_data_overlap, aes(x = PC1, y = PC2, color = Treatment)) +
  geom_point(size = 2.5, alpha = 0.8) +
  stat_ellipse() +
  scale_color_manual(values = col_pal) +
  labs(
    title = "PCA of Overlapping LFMM2 & OutFLANK Outlier SNPs",
    x = paste0("PC1 (", round(pve_overlap[1], 1), "%)"),
    y = paste0("PC2 (", round(pve_overlap[2], 1), "%)")
  ) +
  annotate("text", x = -Inf, y = -Inf,
           label = paste("PERMANOVA p =", format(p_value_overlap, digits = 2)),
           hjust = -0.1, vjust = -1, size = 3.5) +
  theme_classic(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    legend.position = "right"
  )

outlier.pca.overlap.plot
```

# Gene Ontology Enrichment Analysis
Original analysis by Amy Zyck, edits by JBP

## Identifying genes for candidate loci

Using the combined outlier BED file (LFMM2 + OutFLANK) generated above, we intersect with the oyster genome gene annotation to identify candidate genes.

```{bash eval=FALSE}
source activate ROD_CADO
mkdir -p ./analysis/GO
cd ./analysis/GO

# Use the combined outlier BED already generated
bedtools intersect -wb -a ../outlier_detection/results/outliers_lfmm2_outflank_combined.bed -b ../../other_files/sorted.ref3.0.gene.bed | grep -oh "gene=LOC.*;g" | sed 's/gene=//g' | sed 's/;g//g' | sort | uniq > Sig.gene.ROD.CADO.LOC

cat Sig.gene.ROD.CADO.LOC | wc -l
```

## Gene Ontology with TopGO

### Download and prepare GO annotation

```{bash eval=FALSE}
source activate ROD_CADO
cd ./analysis/GO

wget https://ftp.ncbi.nlm.nih.gov/genomes/all/GCF/002/022/765/GCF_002022765.2_C_virginica-3.0/GCF_002022765.2_C_virginica-3.0_gene_ontology.gaf.gz

gunzip GCF_002022765.2_C_virginica-3.0_gene_ontology.gaf.gz

mv GCF_002022765.2_C_virginica-3.0_gene_ontology.gaf gene_ontology.gaf
```

### Create gene-to-GO mapping file

Making a tab file with gene ID and GO IDs. Each gene has one row with all associated GO terms separated by semicolons.

```{bash eval=FALSE}
cd ./analysis/GO

awk 'NR>9 {print $3 "\t" $5}' gene_ontology.gaf | 
sed '1i gene_name\tGeneOntologyIDs' | 
sort | awk '
NR==1 {print; next}
$1!=prev {
if (NR>2) print prev "\t" goids;
prev=$1;
goids=$2;
next
}
{
goids=goids "; " $2
}
END {print prev "\t" goids}
' > gene2go_full.tab
```

### Extract GO terms for candidate genes

```{bash eval=FALSE}
cd ./analysis/GO

grep -F -f Sig.gene.ROD.CADO.LOC gene2go_full.tab > ROD_cand_genes_GO.tab

# Add header
awk 'BEGIN {print "gene_name\tGeneOntologyIDs"} {print}' ROD_cand_genes_GO.tab > ROD_cand_genes_GO_head.tab
```

### Load gene-to-GO data
```{r}
all_gene2go <- read.delim("analysis/GO/gene2go_full.tab", sep="\t")
out_gene2go <- read.delim("analysis/GO/ROD_cand_genes_GO_head.tab", sep="\t")
```

### Prepare gene list for TopGO
```{r}
cand_genes <- as.character(unique(out_gene2go$gene_name))
all_genes <- as.character(unique(all_gene2go$gene_name))
out_GeneList <- factor(as.integer(all_genes %in% cand_genes))
names(out_GeneList) <- all_genes
```

### Read in gene-to-GO mappings
```{r}
gene2go_topgo <- readMappings("analysis/GO/gene2go_full.tab", IDsep=";", sep="\t")
```

### Set gene selection function
```{r}
topDiffGenes <- function(allScore) {
  return(allScore < 0.05)
}
```

### Biological Processes

```{r}
GO_OUT_BP <- new("topGOdata", ontology="BP", gene2GO=gene2go_topgo, allGenes=out_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)
```

```{r}
GO_OUT_BP_FE <- runTest(GO_OUT_BP, algorithm="weight01", statistic="fisher")
```

```{r}
GO_OUT_BP_En <- GenTable(GO_OUT_BP, Fisher = GO_OUT_BP_FE, orderBy = "Fisher", topNodes = 100, numChar = 100)
```

```{r}
GO_OUT_BP_En$Fisher <- as.numeric(GO_OUT_BP_En$Fisher)
GO_OUT_BP_En_sig <- GO_OUT_BP_En[GO_OUT_BP_En$Fisher < 0.05,]
```

```{r}
# Separate GO terms
out_gene2go_sep <- out_gene2go %>%
  separate_rows(GeneOntologyIDs, sep = ";")

out_gene2go_sep$GeneOntologyIDs <- trimws(out_gene2go_sep$GeneOntologyIDs)
GO_OUT_BP_En_sig$GO.ID <- trimws(GO_OUT_BP_En_sig$GO.ID)

GO_OUT_BP_En_sig_gene <- out_gene2go_sep %>%
  left_join(GO_OUT_BP_En_sig, by = c("GeneOntologyIDs" = "GO.ID")) %>%
  na.omit()

GO_OUT_BP_En_sig_gene$ontology <- "Biological Processes"
```

### Cellular Components

```{r}
GO_OUT_CC <- new("topGOdata", ontology="CC", gene2GO=gene2go_topgo, allGenes=out_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)
```

```{r}
GO_OUT_CC_FE <- runTest(GO_OUT_CC, algorithm="weight01", statistic="fisher")
```

```{r}
GO_OUT_CC_En <- GenTable(GO_OUT_CC, Fisher = GO_OUT_CC_FE, orderBy = "Fisher", topNodes = 100, numChar = 100)
```

```{r}
GO_OUT_CC_En$Fisher <- as.numeric(GO_OUT_CC_En$Fisher)
GO_OUT_CC_En_sig <- GO_OUT_CC_En[GO_OUT_CC_En$Fisher < 0.05,]
```

```{r}
GO_OUT_CC_En_sig$GO.ID <- trimws(GO_OUT_CC_En_sig$GO.ID)

GO_OUT_CC_En_sig_gene <- out_gene2go_sep %>%
  left_join(GO_OUT_CC_En_sig, by = c("GeneOntologyIDs" = "GO.ID")) %>%
  na.omit()

GO_OUT_CC_En_sig_gene$ontology <- "Cellular Components"
```

### Molecular Functions

```{r}
GO_OUT_MF <- new("topGOdata", ontology="MF", gene2GO=gene2go_topgo, allGenes=out_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)
```

```{r}
GO_OUT_MF_FE <- runTest(GO_OUT_MF, algorithm="weight01", statistic="fisher")
```

```{r}
GO_OUT_MF_En <- GenTable(GO_OUT_MF, Fisher = GO_OUT_MF_FE, orderBy = "Fisher", topNodes = 100, numChar = 100)
```

```{r}
GO_OUT_MF_En$Fisher <- as.numeric(GO_OUT_MF_En$Fisher)
GO_OUT_MF_En_sig <- GO_OUT_MF_En[GO_OUT_MF_En$Fisher < 0.05,]
```

```{r}
GO_OUT_MF_En_sig$GO.ID <- trimws(GO_OUT_MF_En_sig$GO.ID)

GO_OUT_MF_En_sig_gene <- out_gene2go_sep %>%
  left_join(GO_OUT_MF_En_sig, by = c("GeneOntologyIDs" = "GO.ID")) %>%
  na.omit()

GO_OUT_MF_En_sig_gene$ontology <- "Molecular Functions"
```

### Combine all ontologies

```{r}
GO_OUT_En_sig_gene <- rbind(GO_OUT_BP_En_sig_gene, GO_OUT_CC_En_sig_gene, GO_OUT_MF_En_sig_gene)
```

```{r}
merged_out_genes <- GO_OUT_En_sig_gene %>%
  mutate(prop.sig.genes = Significant/Annotated)

# Save unique instances of gene/GO pairs
merged_out_genes <- unique(merged_out_genes)
```

```{r}
write.csv(merged_out_genes, "analysis/GO/ROD_OUT_GO_en_sig_gene.csv")
```

### Plot GO enrichment

```{r}
GO_plot <- ggplot(merged_out_genes, aes(x = Term, y = -log(Fisher), size = prop.sig.genes, fill = -log(Fisher))) +
  expand_limits(y = 1.5) +
  ylim(1, 7.25) +
  geom_hline(yintercept = -log10(0.01), linetype = "longdash", colour = "black", linewidth = .6) +
  geom_hline(yintercept = -log10(0.001), linetype = "solid", colour = "black", linewidth = .6) +
  geom_point(shape = 21) + 
  scale_size(range = c(2, 12)) + 
  scale_fill_continuous(low = "#1AD3D1FF", high = "#4686FBFF") +
  xlab('') + 
  ylab('Enrichment score') +
  theme_bw() +
  facet_grid(vars(ontology), scales = "free", space = "free_y") +
  coord_flip()

GO_plot
```

```{r}
png(filename = "figures/GO_enrichment_plot.png",
    type = "cairo",
    width = 2000,
    height = 2000,
    res = 200,
    bg = "transparent")

print(GO_plot)
dev.off()
```





