---
title: "Reproducible analysis for Zyck et al. 2026"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(dev = "png", dev.args = list(type = "cairo"))
Sys.setenv(DISPLAY = "")
```


# Setup R and Environment
This scripts assumes the directory structure of the github repository and that all demultiplexed sequencing reads are in the `./analysis` directory

```{r}
library(OutFLANK)
library(bigsnpr)
library(plyr)
library(stringr)
library(reshape2)
library(tidyr)
library(ggplot2)
library(plotrix)
library(ggtext)
library(ggsurvfit)
library(patchwork)
library(pcadapt)
library(LEA)
library(vegan)
library(lfmm)
library(vcfR)
library(adegenet)
library(ade4)
library(tibble)
library(data.table)
library(topGO)
library(dplyr)
```

```{r}
col_pal <- c("#0072B2", "#56B4E9",  "#F0E442", "#E69F00")
theme_set(theme_classic(base_family = "Liberation Sans"))
```
# Mortality Data
```{r}
df.mortality <- read.csv("data/oyster_survival_data.csv", stringsAsFactors = FALSE)
# Keep only the columns needed for survival analysis and clean up
df.mortality <- df.mortality %>%
  select(Treatment, Time, Status) %>%
  filter(!is.na(Time), !is.na(Status)) %>%
  mutate(
    Treatment = factor(
      Treatment,
      levels = c("CONCON", "STRCON", "CONROD", "STRROD")
    ),
    # Status: 1 = event (death), 0 = censored
    Status = as.integer(Status)
  )
df.mortality$Treatment <- factor(df.mortality$Treatment, levels=c("CONCON", "STRCON", "CONROD", "STRROD"))
```

```{r}
km_fit <- survfit2(Surv(Time, Status) ~ Treatment, data = df.mortality)

km.plot <- km_fit %>%
  ggsurvfit(linewidth = 0.9) +
  add_confidence_interval() +
  add_pvalue(
    location   = "annotation",
    x          = 1,
    y          = 0.08,
    size       = 3.5,
    prepend_p  = TRUE
  ) +
  scale_color_manual(
    values = col_pal,
    labels = c("CONCON", "STRCON", "CONROD", "STRROD")
  ) +
  scale_fill_manual(
    values = col_pal,
    labels = c("CONCON", "STRCON", "CONROD", "STRROD")
  ) +
  scale_x_continuous(
    breaks = c(0, 5, 10, 15, 20, 22),
    limits = c(0, 24)
  ) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0, 1)
  ) +
  labs(
    title    = "Kaplan-Meier Survival Curves by Treatment",
    x        = "Days",
    y        = "Cumulative Survival Probability",
    color    = "Treatment",
    fill     = "Treatment"
  ) +
  theme_ggsurvfit_default() +
  theme_classic(base_size = 13)+
  theme(
    plot.title      = element_text(face = "bold", size = 14),
    plot.subtitle   = element_text(size = 11, colour = "grey40"),
    legend.position = "right",
    legend.title    = element_text(face = "bold"),
    axis.title      = element_text(size = 11),
    axis.text       = element_text(size = 10)
  )

km.plot
```
```{r}
df.F.mortality <- df.mortality %>%
  group_by(Treatment) %>%
  summarise(
    Total     = n(),
    Deaths    = sum(Status == 1),
    Mortality = Deaths / Total * 100,
    SE        = sqrt((Mortality/100 * (1 - Mortality/100)) / Total) * 100
  ) %>%
  ungroup()


df.F.mortality$Treatment <- factor(
  df.F.mortality$Treatment,
  levels = c("CONCON", "STRCON", "CONROD", "STRROD")
)

finalmort.plot <- ggplot(df.F.mortality, aes(x = Treatment, y = Mortality, fill = Treatment)) +
  geom_col(width = 0.6, colour = "black", linewidth = 0.4, alpha=0.9) +
  geom_errorbar(aes(ymin = Mortality - SE, ymax = Mortality + SE),width    = 0.2,linewidth = 0.7) +
  scale_fill_manual(values = col_pal) +
  scale_y_continuous(
    limits = c(0, 80),
    breaks = seq(0, 80, 10),
    labels = function(x) paste0(x, "%"),
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(
    title    = "Final Mortality by Treatment Group",
    x        = "Treatment",
    y        = "Mortality (%)"
  ) +
  theme_classic(base_size = 13) +
  theme(
    plot.title      = element_text(face = "bold", size = 14),
    plot.subtitle   = element_text(size = 11, colour = "grey40"),
    axis.title      = element_text(size = 12),
    axis.text       = element_text(size = 11),
    legend.position = "none"
  )


finalmort.plot
```

# Bioinformatics
## SNP calling
```{bash eval=FALSE}
mamba env create -f ROD_CADO.yml
mamba activate ROD_CADO

cd ./analysis

../scripts/dDocent_ngs dDocent.config
```
## SNP filtering
```{bash eval=FALSE}
source activate ROD_CADO

cd ./analysis/raw.vcf
../../scripts/PVCF_filter.sh TRS 36
cd filtered
vcftools --gzvcf SNP.TRS.TRSdp5g75.nDNA.g1.maf01.FIL.vcf.gz --minDP 10 --recode --recode-INFO-all --stdout | vcftools --vcf - --max-missing 1.0 --maf 0.01 --recode --recode-INFO-all --stdout | bcftools view --threads 24 -O z -o SNP.TRSdp10g1.FIL.vcf.gz
bcftools --threads 40 -m2 -M2 -O z -o SNP.TRSdp10g1max2alleles.FIL.vcf.gz SNP.TRSdp10g1.FIL.vcf.gz
plink2 --allow-extra-chr --make-bed --out TRSdp10g1.FIL --vcf SNP.TRSdp10g1max2alleles.FIL.vcf.gz
```

# Diversity Metrics

## Nucleotide diversity
Original analysis by Jacob Green, edits by JBP

### Calculate from vcf files
```{bash, eval=FALSE}
source activate ROD_CADO
mkdir -p ./analysis/diversity_statistics/nd
cd ./analysis/diversity_statistics/nd

cp ../../../treat_popmap .

awk '$2 == "CON-CON" {print $1}' treat_popmap > con-con.txt 
awk '$2 == "STR-CON" {print $1}' treat_popmap > str-con.txt 
awk '$2 == "CON-ROD" {print $1}' treat_popmap > con-rod.txt 
awk '$2 == "STR-ROD" {print $1}' treat_popmap > str-rod.txt

```


```{bash eval=FALSE}
source activate ROD_CADO
cd ./analysis/diversity_statistics/nd


bcftools view --threads 20 ../../raw.vcf/filtered/SNP.TRSdp10g1.FIL.vcf.gz -S con-con.txt | vcftools --vcf - --window-pi 10000 --out con-con

bcftools view --threads 20 ../../raw.vcf/filtered/SNP.TRSdp10g1.FIL.vcf.gz -S str-con.txt | vcftools --vcf - --window-pi 10000 --out str-con

bcftools view --threads 20 ../../raw.vcf/filtered/SNP.TRSdp10g1.FIL.vcf.gz -S con-rod.txt | vcftools --vcf - --window-pi 10000 --out con-rod

bcftools view --threads 20 ../../raw.vcf/filtered/SNP.TRSdp10g1.FIL.vcf.gz -S str-rod.txt | vcftools --vcf - --window-pi 10000 --out str-rod

```

### Load into R
```{r, message=FALSE, warning=FALSE}
pi.concon.dataframe<-read.table("analysis/diversity_statistics/nd/con-con.windowed.pi", sep="\t", header=T)
pi.conrod.dataframe<-read.table("analysis/diversity_statistics/nd/con-rod.windowed.pi", sep="\t", header=T)
pi.strcon.dataframe<-read.table("analysis/diversity_statistics/nd/str-con.windowed.pi", sep="\t", header=T)
pi.strrod.dataframe<-read.table("analysis/diversity_statistics/nd/str-rod.windowed.pi", sep="\t", header=T)

pi.concon.dataframe$Treatment <- "CONCON"
pi.conrod.dataframe$Treatment <- "CONROD"
pi.strcon.dataframe$Treatment <- "STRCON"
pi.strrod.dataframe$Treatment <- "STRROD"

pi.df <- rbind(pi.concon.dataframe,pi.conrod.dataframe,pi.strcon.dataframe, pi.strrod.dataframe)
pi.df$Treatment <- factor(pi.df$Treatment, levels=c("CONCON", "STRCON", "CONROD", "STRROD"))
```

```{r}
pi.plot <- ggplot(pi.df, aes(x=PI,y = Treatment))+
  geom_violin(aes(color=Treatment,fill= Treatment), alpha=0.9) +
  geom_boxplot(width=0.25,outlier.shape = NA, outlier.color = NA, fill = NA)+
  stat_summary(fun=mean, geom="point", shape=23, size=2)+
  scale_fill_manual(values=col_pal)+
  scale_color_manual(values=col_pal)+
  scale_y_discrete(limits = rev(levels(pi.df$Treatment)))+
  labs(x=expression(pi)) +
  ggtitle("Nucleotide Diversity by Treatment")+
  theme_classic(base_size = 13) +
  theme(axis.title.y = element_blank(),plot.title = element_text(face = "bold"),legend.title   = element_text(face = "bold")) 

pi.plot
```
```{r}
group_by(pi.df, Treatment) %>%
  dplyr::summarise(
    count = n(),
    mean = mean(PI, na.rm = TRUE),
    sd = sd(PI, na.rm = TRUE),
    se=std.error(PI, na.rm = TRUE) )
```


```{r}

aov.pi <- aov(PI ~ Treatment, pi.df)
summary(aov.pi)

posthoc.pi <- TukeyHSD(aov.pi)
posthoc.pi


```

## Heterozygosity
Analysis by Coline Caillon with edits by Jon Puritz

```{bash eval=FALSE}
source activate ROD_CADO
mkdir -p ./analysis/diversity_statistics/het
cd ./analysis/diversity_statistics/het

popStats -y GT --file <(bcftools view --threads 40 ../../raw.vcf/filtered/SNP.TRSdp10g1.FIL.vcf.gz) --target 0,1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19 > CONCON.het &

popStats -y GT --file <(bcftools view --threads 40 ../../raw.vcf/filtered/SNP.TRSdp10g1.FIL.vcf.gz) --target 20,21,22,23,24,25,26,27,28,29,30,31,32,33,34,35,36,37,38,39 > STRCON.het &

popStats -y GT --file <(bcftools view --threads 40 ../../raw.vcf/filtered/SNP.TRSdp10g1.FIL.vcf.gz) --target 40,41,42,43,44,45,46,47,48,49,50,51,52,53,54,55,56,57,58,59 > CONROD.het &

popStats -y GT --file <(bcftools view --threads 40 ../../raw.vcf/filtered/SNP.TRSdp10g1.FIL.vcf.gz) --target 60,61,62,63,64,65,66,67,68,69,70,71,72,73,74,75,76,77,78,79 > STRROD.het

```

```{bash}
source activate ROD_CADO
cd ./analysis/diversity_statistics/het

# CON-CON
cat <(echo -e "CHROM\tBIN_START\tBIN_END\tN_SITES\tN_SNPs\tHET\tTREAT") <(mawk '{if (NR>1) print $1 "\t" $2-1 "\t" $2 "\t" $4 "\t" $5 "\tCONCON"}' CONCON.het) > CONCON-treat.het

# STR-CON
cat <(echo -e "CHROM\tBIN_START\tBIN_END\tN_SITES\tN_SNPs\tHET\tTREAT") <(mawk '{if (NR>1) print $1 "\t" $2-1 "\t" $2 "\t" $4 "\t" $5 "\tSTRCON"}' STRCON.het) > STRCON-treat.het

# CON-ROD
cat <(echo -e "CHROM\tBIN_START\tBIN_END\tN_SITES\tN_SNPs\tHET\tTREAT") <(mawk '{if (NR>1) print $1 "\t" $2-1 "\t" $2 "\t" $4 "\t" $5 "\tCONROD"}' CONROD.het) > CONROD-treat.het

# STR-ROD
cat <(echo -e "CHROM\tBIN_START\tBIN_END\tN_SITES\tN_SNPs\tHET\tTREAT") <(mawk '{if (NR>1) print $1 "\t" $2-1 "\t" $2 "\t" $4 "\t" $5 "\tSTRROD"}' STRROD.het) > STRROD-treat.het

```

```{bash}
source activate ROD_CADO
cd ./analysis/diversity_statistics/het
bedtools makewindows -g ../../../other_files/genome.file -w 10000 | mawk '!/NC_007175.2/' > 10kb.bed

bedtools intersect -b CONCON-treat.het -a 10kb.bed -wa -wb | bedtools merge -d -1 -c 7,8 -o mean > CONCON.het.10kb.bed &
bedtools intersect -b CONROD-treat.het -a 10kb.bed -wa -wb | bedtools merge -d -1 -c 7,8 -o mean > CONROD.het.10kb.bed &
bedtools intersect -b STRCON-treat.het -a 10kb.bed -wa -wb | bedtools merge -d -1 -c 7,8 -o mean > STRCON.het.10kb.bed &
bedtools intersect -b STRROD-treat.het -a 10kb.bed -wa -wb | bedtools merge -d -1 -c 7,8 -o mean > STRROD.het.10kb.bed

wait

cat <(echo -e "CHROM\tBIN_START\tBIN_END\tEXP_HET\tOBS_HET\tTREATMENT") <(mawk '{if (NR>1)print $0 "\tCONCON"}' CONCON.het.10kb.bed) <(mawk '{if (NR>1)print $0 "\tCONROD"}' CONROD.het.10kb.bed) <(mawk '{if (NR>1)print $0 "\tSTRCON"}' STRCON.het.10kb.bed) <(mawk '{if (NR>1)print $0 "\tSTRROD"}' STRROD.het.10kb.bed)  > treat.10kb.txt

```

### Load into R
```{r, message=FALSE, warning=FALSE}
het.df <- read.table("analysis/diversity_statistics/het/treat.10kb.txt", sep="\t", header=T)
het.df$Treatment <- factor(het.df$TREATMENT, levels=c("CONCON", "STRCON", "CONROD", "STRROD"))
```

```{r}
het.plot <- ggplot(het.df, aes(x=OBS_HET,y = Treatment))+
  geom_violin(aes(color=Treatment,fill= Treatment), alpha =0.9) +
  geom_boxplot(width=0.25,outlier.shape = NA, outlier.color = NA, fill = NA)+
  stat_summary(fun=mean, geom="point", shape=23, size=2)+
  scale_fill_manual(values=col_pal)+
  scale_color_manual(values=col_pal)+
  scale_y_discrete(limits = rev(levels(het.df$Treatment)))+
  labs(x="Observed Heterozygosity") +
  ggtitle("Observed Heterozygosity by Treatment")+
  theme_classic(base_size = 13) +
    theme(axis.title.y = element_blank(),plot.title = element_text(face = "bold"),legend.title   = element_text(face = "bold")) 
het.plot
```
```{r}
group_by(het.df, Treatment) %>%
  dplyr::summarise(
    count = n(),
    mean = mean(OBS_HET, na.rm = TRUE),
    sd = sd(OBS_HET, na.rm = TRUE),
    se=std.error(OBS_HET) )
```


```{r}

aov.het <- aov(OBS_HET ~ Treatment, het.df)
summary(aov.het)

posthoc.het <- TukeyHSD(aov.het)
posthoc.het


```

## Tajima's D
Code by Megan Guidry with edits by Jon Puritz
```{bash, eval= FALSE}
mamba create -n vkrc python=3.8 numpy scipy
source activate vkrc
pip install VCF-kit
mamba install bcftools
```


```{bash eval=FALSE}
source activate vkrc
mkdir -p ./analysis/diversity_statistics/td
cd ./analysis/diversity_statistics/td

bcftools view --threads 20 ../../raw.vcf/filtered/SNP.TRSdp10g1.FIL.vcf.gz -S ../nd/con-con.txt | vk tajima 10000 10000 - > con-con.tajima &

bcftools view --threads 20 ../../raw.vcf/filtered/SNP.TRSdp10g1.FIL.vcf.gz -S ../nd/str-con.txt | vk tajima 10000 10000 - > str-con.tajima &

bcftools view --threads 20 ../../raw.vcf/filtered/SNP.TRSdp10g1.FIL.vcf.gz -S ../nd/con-rod.txt | vk tajima 10000 10000 - > con-rod.tajima &

bcftools view --threads 20 ../../raw.vcf/filtered/SNP.TRSdp10g1.FIL.vcf.gz -S ../nd/str-rod.txt | vk tajima 10000 10000 - > str-rod.tajima
```


```{bash}
source activate ROD_CADO
cd ./analysis/diversity_statistics/td
cat <(echo -e "CHROM\tBIN_START\tBIN_END\tN_SITES\tN_SNPs\tTAJD\tTREATMENT") <(mawk '{if (NR>1)print $0 "\tCONCON"}' con-con.tajima) <(mawk '{if (NR>1)print $0 "\tSTRCON"}' str-con.tajima) <(mawk '{if (NR>1)print $0 "\tCONROD"}' con-rod.tajima) <(mawk '{if (NR>1)print $0 "\tSTRROD"}' str-rod.tajima) > treatment.tajima
```

### Load into R
```{r, message=FALSE, warning=FALSE}
tajd.df <- read.table("analysis/diversity_statistics/td/treatment.tajima", sep="\t", header=T)
tajd.df$Treatment <- factor(tajd.df$TREATMENT, levels=c("CONCON", "STRCON", "CONROD", "STRROD"))
```

```{r}
tajd.plot <- ggplot(tajd.df, aes(x=TAJD,y = Treatment))+
  geom_violin(aes(color=Treatment,fill= Treatment)) +
  geom_boxplot(width=0.25,outlier.shape = NA, outlier.color = NA, fill = NA)+
  stat_summary(fun=mean, geom="point", shape=23, size=2)+
  scale_fill_manual(values=col_pal)+
  scale_color_manual(values=col_pal)+
  scale_y_discrete(limits = rev(levels(tajd.df$Treatment)))+
  labs(x = expression("Tajima's "*italic(D))) +
  ggtitle(expression("Tajima's "*italic(D)*" by Treatment")) +
  theme_classic(base_size = 13) +
  theme(axis.title.y = element_blank(), text = element_text(family = "Liberation Sans"))

tajd.plot


```

```{r}
group_by(tajd.df, Treatment) %>%
  dplyr::summarise(
    count = n(),
    mean = mean(TAJD, na.rm = TRUE),
    sd = sd(TAJD, na.rm = TRUE),
    se=std.error(TAJD) )
```

```{r}

aov.tajd <- aov(TAJD ~ Treatment, tajd.df)
summary(aov.tajd)

posthoc.tajd <- TukeyHSD(aov.tajd)
posthoc.tajd

```

# PCA
Original Code by Jacob Green, edits by Jon Puritz
```{bash, eval= FALSE}
cd analysis/raw.vcf/filtered/
plink2 --bfile TRSdp10g1.FIL --make-bed --allow-extra-chr --out TRSdp10g1.FIL.ALL --maf 0.05


```

```{r}
#Set path to file
path_to_file <- "analysis/raw.vcf/filtered/TRSdp10g1.FIL.bed"
#Load file to filename variable
filename <- read.pcadapt(path_to_file, type = "bed")

resld <- pcadapt(filename, K = 10, LD.clumping = list(size = 500, thr = 0.2))
plot(resld, option = "screeplot")
```

```{r}
poplist.names <- c(rep("CONCON",20),rep("STRCON",20),rep("CONROD",20),rep("STRROD",20))
poplist.names
```

```{r}
df.pca <- as.data.frame(resld$scores)
colnames(df.pca) <- paste0("PC", 1:ncol(df.pca))

df.pca$Treatment <- poplist.names
df.pca$Treatment <- factor(df.pca$Treatment, levels=c("CONCON", "STRCON", "CONROD", "STRROD"))

variance_explained <- resld$singular.values^2 / sum(resld$singular.values^2) * 100
cat("Variance explained by each PC:\n")
print(round(variance_explained, 2))

pca.plot <- ggplot(df.pca, aes(x = PC1, y = PC2, colour = Treatment)) +
  geom_point(size = 2.5, alpha = 0.85) +
  labs(
    title = "PCA of Genetic Variation",
    x = paste0("PC1 (", round(variance_explained[1], 1), "%)"),
    y = paste0("PC2 (", round(variance_explained[2], 1), "%)")
  ) +
  theme_classic(base_size = 13) +
  scale_fill_manual(values=col_pal)+
  scale_color_manual(values=col_pal)+
  theme(
    plot.title     = element_text(face = "bold"),
    legend.title   = element_text(face = "bold"),
    legend.position = "right"
  )

pca.plot
```

# Figure 1
```{bash}
mkdir -p figures
```

```{r}
png(filename="figures/Figure1.png", type="cairo",units="px", width=5600, height=3000, res=300, bg="transparent")

gen_div.plot <- (pi.plot / het.plot)

(finalmort.plot + pca.plot)/(km.plot + gen_div.plot) + plot_layout(guides="collect") + plot_annotation(tag_levels = 'A')
dev.off()
```

# Outlier Detection
Original analysis by Amy Zyck, edits by JBP

## Setup

### Prepare genotype file for LFMM2
```{bash eval=FALSE}
source activate ROD_CADO
mkdir -p ./analysis/outlier_detection
cd ./analysis/outlier_detection
pwd

plink2 --vcf ../raw.vcf/filtered/SNP.TRSdp10g1.FIL.vcf.gz --allow-extra-chr --min-alleles 2 --max-alleles 2 --export A --out snp_rod
cut -f7- snp_rod.raw | tail -n +2 | sed 's/NA/9/g' > snp_rod.lfmm
rm snp_rod.raw
```

### Prepare environmental strata
```{bash eval=FALSE}
cd ./analysis/outlier_detection
cp -f ../../other_files/strata_ROD_tab .
```

### Create SNP index file
```{bash eval=FALSE}
source activate ROD_CADO
cd ./analysis/outlier_detection

bcftools view -m 2 -M 2 ../raw.vcf/filtered/SNP.TRSdp10g1.FIL.vcf.gz | bcftools query -f '%CHROM\t%POS\n'  > totalloci
NUM=(`cat totalloci | wc -l`)
paste <(seq 1 $NUM) totalloci > loci.plus.index
```

## Run LFMM2


### Load genotype and environmental data
```{r}
clim.env_rod <- read.table("analysis/outlier_detection/strata_ROD_tab", header=TRUE)
strata_ROD <- clim.env_rod
gen_rod <- fread("analysis/outlier_detection/snp_rod.lfmm", header = FALSE)
row.names(gen_rod) <- clim.env_rod$Individual

CADO <- clim.env_rod$CADO
DIS <- clim.env_rod$Disease
COMB_BIN <- clim.env_rod$Combined_Bin
```

### LFMM ridge - Stress (CADO)
```{r eval=FALSE}
oys_rod_CADO.lfmm <- lfmm_ridge(Y=gen_rod, X=CADO, K=4)
oys_rod_CADO.pv <- lfmm_test(Y=gen_rod, X=CADO, lfmm=oys_rod_CADO.lfmm, calibrate="gif")
oys_rod_CADO.pv$gif
```

```{r eval= FALSE}
save(oys_rod_CADO.lfmm, oys_rod_CADO.pv, file = "./analysis/outlier_detection/lfmm1_data.RData")
```

```{r}
load("./analysis/outlier_detection/lfmm1_data.RData")

```


```{r}
zscore_rod_CADO <- oys_rod_CADO.pv$score[,1]
adj.pv_rod_CADO <- pchisq(zscore_rod_CADO^2/0.98, df=1, lower = FALSE)
oys_rod_CADO.qv <- qvalue(adj.pv_rod_CADO)$qvalues
oys_rod_CADO.FDR.1 <- as.vector(which(oys_rod_CADO.qv < 0.1))
length(oys_rod_CADO.FDR.1)
```

```{r}
cat("", file = "analysis/outlier_detection/outliers_lfmm2_CADO_098.txt")
invisible(lapply(oys_rod_CADO.FDR.1, write, "analysis/outlier_detection/outliers_lfmm2_CADO_098.txt", append=TRUE))
```

### LFMM ridge - Disease
```{r eval=FALSE}
oys_rod_DIS.lfmm <- lfmm_ridge(Y=gen_rod, X=DIS, K=4)
oys_rod_DIS.pv <- lfmm_test(Y=gen_rod, X=DIS, lfmm=oys_rod_DIS.lfmm, calibrate="gif")
oys_rod_DIS.pv$gif
```

```{r eval= FALSE}
save(oys_rod_DIS.lfmm, oys_rod_DIS.pv, file = "./analysis/outlier_detection/lfmm2_data.RData")
```

```{r}
load("./analysis/outlier_detection/lfmm2_data.RData")

```


```{r}
zscore_rod_DIS <- oys_rod_DIS.pv$score[,1]
adj.pv_rod_DIS <- pchisq(zscore_rod_DIS^2/0.98, df=1, lower = FALSE)
oys_rod_DIS.qv <- qvalue(adj.pv_rod_DIS)$qvalues
oys_rod_DIS.FDR.1 <- as.vector(which(oys_rod_DIS.qv < 0.1))
length(oys_rod_DIS.FDR.1)
```

```{r}
cat("", file = "analysis/outlier_detection/outliers_lfmm2_DIS_098.txt")
invisible(lapply(oys_rod_DIS.FDR.1, write, "analysis/outlier_detection/outliers_lfmm2_DIS_098.txt", append=TRUE))
```

### LFMM ridge - Combined treatment (binary)
```{r eval=FALSE}
oys_rod_CB.lfmm <- lfmm_ridge(Y=gen_rod, X=COMB_BIN, K=4)
oys_rod_CB.pv <- lfmm_test(Y=gen_rod, X=COMB_BIN, lfmm=oys_rod_CB.lfmm, calibrate="gif")
oys_rod_CB.pv$gif
```

```{r eval= FALSE}
save(oys_rod_CB.lfmm, oys_rod_CB.pv, file = "./analysis/outlier_detection/lfmm3_data.RData")
```

```{r}
load("./analysis/outlier_detection/lfmm3_data.RData")

```


```{r}
oys_rod_CB.qv <- qvalue(oys_rod_CB.pv$calibrated.pvalue)$qvalues
oys_rod_CB.FDR.1 <- which(oys_rod_CB.qv < 0.1)
length(oys_rod_CB.FDR.1)
```

```{r}
cat("", file = "analysis/outlier_detection/outliers_lfmm2_CB.txt")
invisible(lapply(oys_rod_CB.FDR.1, write, "analysis/outlier_detection/outliers_lfmm2_CB.txt", append=TRUE))
```

## Combine outliers and create BED file

### Map SNP indices to genomic positions
```{bash eval=TRUE}
source activate ROD_CADO
mkdir -p analysis/outlier_detection/results
cd ./analysis/outlier_detection

cat outliers_lfmm2_CADO_098.txt | parallel "grep -w ^{} loci.plus.index" | cut -f2,3 > ./results/outliers_lfmm2_CADO_098.loci.txt
cat outliers_lfmm2_DIS_098.txt | parallel "grep -w ^{} loci.plus.index" | cut -f2,3 > ./results/outliers_lfmm2_DIS_098.loci.txt
cat outliers_lfmm2_CB.txt | parallel "grep -w ^{} loci.plus.index" | cut -f2,3 > ./results/outliers_lfmm2_CB.loci.txt

cat ./results/outliers_lfmm2_CB.loci.txt ./results/outliers_lfmm2_CADO_098.loci.txt ./results/outliers_lfmm2_DIS_098.loci.txt > ./results/all_lfmm2_outliers
awk '{print $1 "\t" $2-1 "\t" $2}' ./results/all_lfmm2_outliers | bedtools sort -i - > ./results/outliers_lfmm2.bed
```

### Create outlier VCF for validation
```{bash eval=FALSE}
source activate ROD_CADO
cd ./analysis/outlier_detection

bcftools index --threads 20 ../raw.vcf/filtered/SNP.TRSdp10g1.FIL.vcf.gz

bcftools view --threads 20 -R ./results/outliers_lfmm2_CB.loci.txt ../raw.vcf/filtered/SNP.TRSdp10g1.FIL.vcf.gz -O b -o outliers_lfmm2_CB.recode.vcf
```

# FST and OutFLANK
```{bash, eval= FALSE}
source activate ROD_CADO
mkdir -p ./analysis/FST
cd ./analysis/FST

plink2 --bfile ../raw.vcf/filtered/TRSdp10g1.FIL --make-bed --allow-extra-chr --out TRSdp10g1.FIL.ALL --maf 0.05

```


```{r}
# Read in bedfile
rds <- snp_readBed("./analysis/raw.vcf/filtered/TRSdp10g1.FIL.ALL.bed", backingfile = tempfile())
```

```{r, eval = FALSE}
full_bed <- snp_attach(rds)
G <- full_bed$genotypes
CHR <- full_bed$map$chromosome
POS <- full_bed$map$physical.pos
NCORES <- nb_cores()
```

```{r, eval=FALSE}
svd <- snp_autoSVD(G, as.integer(factor(CHR)), POS, ncores = NCORES,min.mac =8,size=10)
```

```{r, eval=FALSE}
write(paste(CHR[attr(svd, "subset")],
            POS[attr(svd, "subset")], sep='\t'),
      "./analysis/FST/ThinnedSNPs.maskedPCs.txt")
```

```{bash, eval= FALSE}
source activate ROD_CADO
cd ./analysis/FST
rm -f random.50k.snps

cut -f1 ThinnedSNPs.maskedPCs.txt| uniq | parallel "grep {} ThinnedSNPs.maskedPCs.txt | shuf  2>/dev/null | shuf 2>/dev/null | shuf 2>/dev/null| head -5000" >> random.50k.snps

```

```{bash, eval=FALSE}
source activate ROD_CADO
cd ./analysis/FST

plink2 --bfile ../raw.vcf/filtered/TRSdp10g1.FIL --make-bed --allow-extra-chr --extract bed0 <(mawk '{print $1"\t"$2-1"\t"$2}' random.50k.snps ) --out random50k

mawk 'NR==1{print "IND\tPOP"};NR >0 {print $0}' ../popmap > popmap.txt
```

```{r}
rds.random <- snp_readBed("./analysis/FST/random50k.bed", backingfile = tempfile())
popmap <- read.table("./analysis/FST/popmap.txt", header=TRUE)
popmap$TREAT <- c(rep("CONCON",20),rep("STRCON",20),rep("CONROD",20),rep("STRROD",20))

random <- snp_attach(rds.random)
G.r <- random$genotypes
CHR.r <- random$map$chromosome
POS.r <- random$map$physical.pos

```

```{r eval=FALSE}
my_fst <- MakeDiploidFSTMat(G[], locusNames = paste0(CHR,"_", POS), popNames = popmap$TREAT)

my_fst.r <- MakeDiploidFSTMat(G.r[], locusNames = paste0(CHR.r,"_", POS.r), popNames = popmap$TREAT)
```




```{r eval=FALSE}
save(my_fst, file = "./analysis/FST/FST_data.RData")
```

```{r}
load("./analysis/FST/FST_data.RData")

```


```{r}

out_trim <- OutFLANK(my_fst[attr(svd, which="subset"),], NumberOfSamples=4, qthreshold = 0.1, Hmin = 0.1)

OutFLANKResultsPlotter(out_trim, withOutliers = TRUE,
                       NoCorr = TRUE, Hmin = 0.1, binwidth = 0.001, Zoom =
                         FALSE, RightZoomFraction = 0.05, titletext = NULL)

hist(out_trim$results$pvaluesRightTail)

qthres <- 0.1

P1 <- pOutlierFinderChiSqNoCorr(my_fst, Fstbar = out_trim$FSTNoCorrbar, 
                                   dfInferred = out_trim$dfInferred, qthreshold = qthres, Hmin=0.1)

P1 <- P1[order(as.numeric(rownames(P1))),,drop=FALSE]                  
hist(P1$pvaluesRightTail)
```


```{r}
plot.df<- setNames(data.frame(matrix(ncol = 8, nrow = length(P1$LocusName))), c("CHR", "CHRR", "BP", "SNP", "P", "F", "Q", "Outlier"))
plot.df$CHR <-as.integer(as.factor(CHR))
plot.df$CHRR <- CHR
plot.df$SNP <-P1$LocusName
plot.df$BP <- POS
plot.df$P <- P1$pvalues
plot.df$Q <- P1$qvalues
plot.df$F <- P1$FST
plot.df$Outlier <- P1$OutlierFlag
plot.df <- plot.df[which(plot.df$Outlier == TRUE | plot.df$Outlier == FALSE),]


```

```{r}
withr::with_options(c(scipen = 10), write(paste(plot.df$CHRR,plot.df$BP-1,plot.df$BP,plot.df$F,plot.df$Outlier, sep='\t'), "./analysis/FST/FST_Outflank.bed"))
```

```{bash}
source activate ROD_CADO
cd ./analysis/FST
bedtools intersect -b <(sort -k 1,1 -k2,2n FST_Outflank.bed) -a <(sort -k 1,1 -k2,2n ../diversity_statistics/het/10kb.bed) -wa -wb -sorted  |bedtools merge -d -1 -c 7,8 -o mean,distinct | mawk 'BEGIN {print "CHROM\tBIN_START\tBIN_END\tWEIGHTED_FST\tOUTLIER"} { if ($0 ~ /TRUE/) {print $1 "\t" $2 "\t" $3 "\t" $4 "\tTRUE"} else {print $0}}' > OF_fst.10kb.bed

cat <(echo -e "CHROM\tBIN_START\tBIN_END\tWEIGHTED_FST\tOUTLIER") OF_fst.10kb.bed <(grep TRUE FST_Outflank.bed) > OF_fst.10kb.txt



cat <(echo -e "CHROM\tBIN_START\tBIN_END\tWEIGHTED_FST\tOUTLIER") FST_Outflank.bed > outlier_fst.txt
```


```{r}
fst <- fread("./analysis/FST/outlier_fst.txt", header = TRUE)
SNP<-c(1:(nrow(fst)))
mydf<-data.frame(SNP,fst)
mydf$CHR <- mydf$CHROM
mydf %>% 
  mutate(CHR = str_replace(CHR, "NC_035780.1", "1")) %>% 
  mutate(CHR = str_replace(CHR, "NC_035781.1", "2")) %>% 
  mutate(CHR = str_replace(CHR, "NC_035782.1", "3")) %>% 
  mutate(CHR = str_replace(CHR, "NC_035783.1", "4")) %>% 
  mutate(CHR = str_replace(CHR, "NC_035784.1", "5")) %>% 
  mutate(CHR = str_replace(CHR, "NC_035785.1", "6")) %>% 
  mutate(CHR = str_replace(CHR, "NC_035786.1", "7")) %>% 
  mutate(CHR = str_replace(CHR, "NC_035787.1", "8")) %>%
  mutate(CHR = str_replace(CHR, "NC_035788.1", "9")) %>% 
  mutate(CHR = str_replace(CHR, "NC_035789.1", "10"))  -> mydf
mydf$CHR <- as.numeric(mydf$CHR)  

# Step 1: Compute chromosome-level metrics
chr_offsets <- mydf %>%
  group_by(CHR) %>%
  summarize(
    minbp = min(BIN_START, na.rm = TRUE),
    maxbp = max(BIN_END, na.rm = TRUE),
    nrows = n()
  ) %>%
  mutate(
    offset = cumsum(nrows) - nrows,
    bp_range = ifelse(maxbp == minbp, 1, maxbp - minbp)
  )

# Step 2: Join offset info into the original dataframe and compute BPcum
don <- mydf %>%
  left_join(chr_offsets, by = "CHR") %>%
  mutate(
    scaled_bp = (as.numeric(BIN_START) - as.numeric(minbp)) * as.numeric(nrows) / as.numeric(bp_range),
    BPcum = scaled_bp + as.numeric(offset)
  )

# Step 3: Axis label positions
axisdf <- don %>%
  group_by(CHR) %>%
  summarize(center = mean(BPcum, na.rm = TRUE))




m2<- ggplot(don, aes(x=BPcum, y=WEIGHTED_FST)) +
  geom_point( aes(color=as.factor(CHR)), alpha=0.25, size=0.5) +
  geom_point(data=subset(don, OUTLIER==TRUE),aes(color=as.factor(CHR)),shape = 17,alpha=1,size=1.75) +
  scale_color_manual(values = rep(c("#7d5577", "#52374e"), 5 )) +
  scale_x_continuous( label = axisdf$CHR, breaks= axisdf$center ,expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0),limits=c(min(don$WEIGHTED_FST-0.001,na.rm=TRUE),0.5))+      
  theme(axis.title.y = element_markdown())+ 
  guides(colour = "none",alpha="none", fill="none") +
  labs(x = "Chromosome", y = "*F<sub>ST</sub>*") +
  theme_classic() +
  theme(axis.title.y = element_markdown()) 
    

png(filename="figures/Figure.X.FST.man.png", type="cairo",units="px", width=5600, height=3000, res=300, bg="transparent")
m2
null <- dev.off()

```

# Outlier Overlap and Annotation
Original analysis by Amy Zyck, edits by JBP

## Extract OutFLANK outliers to BED format
```{bash eval=FALSE}
source activate ROD_CADO
cd ./analysis/FST

awk '$5=="TRUE"{print $1"\t"$2"\t"$3}' FST_Outflank.bed > Outflank_outliers.bed
```

## Compare overlapping outlier SNPs between LFMM2 and OutFLANK

```{bash eval=FALSE}
source activate ROD_CADO
cd ./analysis

# Sort each bed file
sort ./FST/Outflank_outliers.bed > ./FST/outliers_outflank_sorted.txt
sort ./outlier_detection/results/outliers_lfmm2.bed > ./outlier_detection/results/outliers_lfmm2_sorted.txt
```

```{bash eval=FALSE}
cd ./analysis

# Shared outlier SNPs between OutFLANK and LFMM2
comm -12 ./FST/outliers_outflank_sorted.txt ./outlier_detection/results/outliers_lfmm2_sorted.txt | wc -l
comm -12 ./FST/outliers_outflank_sorted.txt ./outlier_detection/results/outliers_lfmm2_sorted.txt
```

## Validate all LFMM2 outliers with PCA and PERMANOVA

### Create VCF for all LFMM2 outliers
```{bash eval=FALSE}
source activate ROD_CADO
cd ./analysis/outlier_detection

bcftools view --threads 20 -R ./results/outliers_lfmm2.bed ../raw.vcf/filtered/SNP.TRSdp10g1.FIL.vcf.gz > outliers_lfmm2_all.recode.vcf

```

### Load and analyze all LFMM2 outliers
```{r}
vcf_out_lfmm_all <- read.vcfR("analysis/outlier_detection/outliers_lfmm2_all.recode.vcf")
rad_out_lfmm_all.filt <- vcfR2genind(vcf_out_lfmm_all, strata = strata_ROD, pop = popmap$TREAT)
```

```{r}
x_out_lfmm_all <- scaleGen(rad_out_lfmm_all.filt, NA.method = "mean")
out_lfmm_all_pca <- dudi.pca(x_out_lfmm_all, center = FALSE, scale = FALSE, scannf = FALSE, nf = 4)
pc_scores_lfmm_all <- out_lfmm_all_pca$li
```

```{r}
permanova_result_lfmm_all <- adonis2(pc_scores_lfmm_all ~ poplist.names, method = "euclidean")
print(permanova_result_lfmm_all)
```

```{r}
pca_data_lfmm_all <- data.frame(
  PC1 = pc_scores_lfmm_all[,1],
  PC2 = pc_scores_lfmm_all[,2],
  Treatment = factor(poplist.names, levels = c("CONCON", "STRCON", "CONROD", "STRROD"))
)

p_value_lfmm_all <- permanova_result_lfmm_all$`Pr(>F)`[1]
eigenvalues_lfmm_all <- eigen(cov(pca_data_lfmm_all[, c("PC1", "PC2")]))$values
pve_lfmm_all <- eigenvalues_lfmm_all / sum(eigenvalues_lfmm_all) * 100

outlier.pca.lfmm.all.plot <- ggplot(pca_data_lfmm_all, aes(x = PC1, y = PC2, color = Treatment)) +
  geom_point(size = 2.5, alpha = 0.8) +
  stat_ellipse() +
  scale_color_manual(values = col_pal) +
  labs(
    title = "PCA of All LFMM2 Outlier SNPs",
    x = paste0("PC1 (", round(pve_lfmm_all[1], 1), "%)"),
    y = paste0("PC2 (", round(pve_lfmm_all[2], 1), "%)")
  ) +
  annotate("text", x = -Inf, y = -Inf,
           label = paste("PERMANOVA p =", format(p_value_lfmm_all, digits = 2)),
           hjust = -0.1, vjust = -1, size = 3.5) +
  theme_classic(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    legend.position = "right"
  )

outlier.pca.lfmm.all.plot
```

## Validate OutFLANK outliers with PCA and PERMANOVA

### Create VCF for OutFLANK outliers
```{bash eval=FALSE}
source activate ROD_CADO
cd ./analysis/FST

bcftools view --threads 20 -R Outflank_outliers.bed ../raw.vcf/filtered/SNP.TRSdp10g1.FIL.vcf.gz > outliers_Outflank.recode.vcf
```

### Load and analyze OutFLANK outliers
```{r}
vcf_out_outflank <- read.vcfR("analysis/FST/outliers_Outflank.recode.vcf")
rad_out_outflank.filt <- vcfR2genind(vcf_out_outflank, strata = strata_ROD, pop = popmap$TREAT)
```

```{r}
x_out_outflank <- scaleGen(rad_out_outflank.filt, NA.method = "mean")
out_outflank_pca <- dudi.pca(x_out_outflank, center = FALSE, scale = FALSE, scannf = FALSE, nf = 4)
pc_scores_outflank <- out_outflank_pca$li
```

```{r}
permanova_result_outflank <- adonis2(pc_scores_outflank ~ popmap$TREAT, method = "euclidean")
print(permanova_result_outflank)
```

```{r}
pca_data_outflank <- data.frame(
  PC1 = pc_scores_outflank[,1],
  PC2 = pc_scores_outflank[,2],
  Treatment = factor(poplist.names, levels = c("CONCON", "STRCON", "CONROD", "STRROD"))
)

p_value_outflank <- permanova_result_outflank$`Pr(>F)`[1]
eigenvalues_outflank <- eigen(cov(pca_data_outflank[, c("PC1", "PC2")]))$values
pve_outflank <- eigenvalues_outflank / sum(eigenvalues_outflank) * 100

outlier.pca.outflank.plot <- ggplot(pca_data_outflank, aes(x = PC1, y = PC2, color = Treatment)) +
  geom_point(size = 2.5, alpha = 0.8) +
  stat_ellipse() +
  scale_color_manual(values = col_pal) +
  labs(
    title = "PCA of OutFLANK Outlier SNPs",
    x = paste0("PC1 (", round(pve_outflank[1], 1), "%)"),
    y = paste0("PC2 (", round(pve_outflank[2], 1), "%)")
  ) +
  annotate("text", x = -Inf, y = -Inf,
           label = paste("PERMANOVA p =", format(p_value_outflank, digits = 2)),
           hjust = -0.1, vjust = -1, size = 3.5) +
  theme_classic(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    legend.position = "right"
  )

outlier.pca.outflank.plot
```

## PCA of combined LFMM2 and OutFLANK outliers

### Create combined outlier BED and VCF
```{bash eval=FALSE}
source activate ROD_CADO
cd ./analysis

# Merge LFMM2 and OutFLANK outlier BEDs, sort and deduplicate
cat ./outlier_detection/results/outliers_lfmm2.bed ./FST/Outflank_outliers.bed | bedtools sort -i - | bedtools merge -i - > ./outlier_detection/results/outliers_lfmm2_outflank_combined.bed

bcftools view --threads 20 -R ./outlier_detection/results/outliers_lfmm2_outflank_combined.bed ./raw.vcf/filtered/SNP.TRSdp10g1.FIL.vcf.gz -O v -o ./outlier_detection/outliers_combined.recode.vcf
```

### Load and analyze combined outliers
```{r}
vcf_out_combined <- read.vcfR("analysis/outlier_detection/outliers_combined.recode.vcf")
rad_out_combined.filt <- vcfR2genind(vcf_out_combined, strata = strata_ROD, pop = popmap$TREAT)
```

```{r}
x_out_combined <- scaleGen(rad_out_combined.filt, NA.method = "mean")
out_combined_pca <- dudi.pca(x_out_combined, center = FALSE, scale = FALSE, scannf = FALSE, nf = 4)
pc_scores_combined <- out_combined_pca$li
```

```{r}
permanova_result_combined <- adonis2(pc_scores_combined ~ popmap$TREAT, method = "euclidean")
print(permanova_result_combined)
```

```{r}
pca_data_combined <- data.frame(
  PC1 = pc_scores_combined[,1],
  PC2 = pc_scores_combined[,2],
  Treatment = factor(poplist.names, levels = c("CONCON", "STRCON", "CONROD", "STRROD"))
)

p_value_combined <- permanova_result_combined$`Pr(>F)`[1]
eigenvalues_combined <- eigen(cov(pca_data_combined[, c("PC1", "PC2")]))$values
pve_combined <- eigenvalues_combined / sum(eigenvalues_combined) * 100

outlier.pca.combined.plot <- ggplot(pca_data_combined, aes(x = PC1, y = PC2, color = Treatment)) +
  geom_point(size = 2.5, alpha = 0.8) +
  stat_ellipse() +
  scale_color_manual(values = col_pal) +
  labs(
    title = "PCA of Combined LFMM2 + OutFLANK Outlier SNPs",
    x = paste0("PC1 (", round(pve_combined[1], 1), "%)"),
    y = paste0("PC2 (", round(pve_combined[2], 1), "%)")
  ) +
  annotate("text", x = -Inf, y = -Inf,
           label = paste("PERMANOVA p =", format(p_value_combined, digits = 2)),
           hjust = -0.1, vjust = -1, size = 3.5) +
  theme_classic(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    legend.position = "right"
  )

outlier.pca.combined.plot
```

## PCA of overlapping LFMM2 and OutFLANK outlier loci

### Create overlapping outlier BED and VCF
```{bash eval=FALSE}
source activate ROD_CADO
cd ./analysis

# Extract loci shared between LFMM2 and OutFLANK
bedtools intersect -a ./outlier_detection/results/outliers_lfmm2.bed -b ./FST/Outflank_outliers.bed > ./outlier_detection/results/outliers_overlap_lfmm2_outflank.bed

wc -l ./outlier_detection/results/outliers_overlap_lfmm2_outflank.bed

bcftools view --threads 20 -R ./outlier_detection/results/outliers_overlap_lfmm2_outflank.bed ./raw.vcf/filtered/SNP.TRSdp10g1.FIL.vcf.gz -O v -o ./outlier_detection/outliers_overlap.recode.vcf
```

### Load and analyze overlapping outliers
```{r}
vcf_out_overlap <- read.vcfR("analysis/outlier_detection/outliers_overlap.recode.vcf")
rad_out_overlap.filt <- vcfR2genind(vcf_out_overlap, strata = strata_ROD, pop = popmap$TREAT)
```

```{r}
x_out_overlap <- scaleGen(rad_out_overlap.filt, NA.method = "mean")
out_overlap_pca <- dudi.pca(x_out_overlap, center = FALSE, scale = FALSE, scannf = FALSE, nf = 4)
pc_scores_overlap <- out_overlap_pca$li
```

```{r}
permanova_result_overlap <- adonis2(pc_scores_overlap ~ popmap$TREAT, method = "euclidean")
print(permanova_result_overlap)
```

```{r}
pca_data_overlap <- data.frame(
  PC1 = pc_scores_overlap[,1],
  PC2 = pc_scores_overlap[,2],
  Treatment = factor(poplist.names, levels = c("CONCON", "STRCON", "CONROD", "STRROD"))
)

p_value_overlap <- permanova_result_overlap$`Pr(>F)`[1]
eigenvalues_overlap <- eigen(cov(pca_data_overlap[, c("PC1", "PC2")]))$values
pve_overlap <- eigenvalues_overlap / sum(eigenvalues_overlap) * 100

outlier.pca.overlap.plot <- ggplot(pca_data_overlap, aes(x = PC1, y = PC2, color = Treatment)) +
  geom_point(size = 2.5, alpha = 0.8) +
  stat_ellipse() +
  scale_color_manual(values = col_pal) +
  labs(
    title = "PCA of Overlapping LFMM2 & OutFLANK Outlier SNPs",
    x = paste0("PC1 (", round(pve_overlap[1], 1), "%)"),
    y = paste0("PC2 (", round(pve_overlap[2], 1), "%)")
  ) +
  annotate("text", x = -Inf, y = -Inf,
           label = paste("PERMANOVA p =", format(p_value_overlap, digits = 2)),
           hjust = -0.1, vjust = -1, size = 3.5) +
  theme_classic(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    legend.position = "right"
  )

outlier.pca.overlap.plot
```
# Figure 2 — Genome-wide LFMM association and focal outlier regions

Panel A is the genome-wide LFMM Manhattan; panel B is a grid of focal outlier regions
showing the per-SNP association signal, π, observed heterozygosity and Tajima's *D* by
treatment. The outlier PCA is optional (`show.pca`).

Focal loci can be chosen two ways (`focal.by`):

* `"pvalue"` — the most significant SNPs, thinned by distance. Purely association-driven,
  fully reproducible from this document alone.
* `"file"` — read an explicit list from `analysis/figure2/focal_loci.csv` (columns `CHROM`,
  `POS`, optional `name`). Use this to plot a hand-chosen or externally ranked set.

The STRROD-separation ranking that produced one such list lives outside this document, in
`exploratory/separation_score.R`, and is not tracked in the repository. It is kept separate
deliberately: it ranks loci on the same statistics these panels then display, so separation
visible in the panels is a consequence of the selection, not evidence for it. Panels chosen
that way are exemplars — any genome-wide claim rests on the outlier test and the
treatment-level statistics, not on these columns.

Because LD decays over hundreds of bp to a few kb in *C. virginica*, the panels use a ±20 kb
window and re-bin the diversity metrics at 500 bp. The genome-wide 10 kb bins used elsewhere
are far too coarse at this scale.

Depends on objects built above: the LFMM fits (`oys_rod_CB.pv`, `adj.pv_rod_CADO`,
`adj.pv_rod_DIS` and their q-values), and — only when `show.pca = TRUE` —
`pca_data_combined`, `pve_combined`, `p_value_combined`.

## Settings

`flank` and `binsize` require re-running the bash chunk. Everything else is R-side only.

```{r}
lfmm.var      <- "Combined"   # "Combined" | "CADO" | "Disease" — variable shown in panel A
focal.by      <- "file"       # "file" = analysis/figure2/focal_loci.csv | "pvalue" = top LFMM hits
n.focal       <- 6            # max panels in "pvalue" mode (ignored in "file" mode)
min.sep       <- 50000        # bp; keeps two panels from showing the same locus twice
merge.within  <- 1000         # bp; focal SNPs closer than this share one panel (0 = never)
genic.only    <- FALSE        # TRUE skips SNPs with no overlapping gene (pvalue mode)
flank         <- 15000        # bp either side of the focal SNP
binsize       <- 500          # bp, for the focal-region diversity tracks
smooth.bins   <- 8            # centred rolling mean over this many bins; 1 = raw
show.pca      <- FALSE        # TRUE puts the outlier PCA back in as a panel
ins.top       <- "FST"       # top row of the focal panels: "LFMM" or "FST"
strip.theme   <- NA           # show the GO theme on the panel strip; NA = always (3-row label)

# --- panel A rendering speed -------------------------------------------------
# The dominant cost is that every inline chunk preview redraws the millions of points.
# With preview = FALSE the panels are built but not printed, so the Manhattan is drawn
# once (into the PNG) instead of three times. Set TRUE while iterating on panel design.
preview <- FALSE

# Truncate panel A: draw only SNPs with p <= man.pmax. Unlike random thinning this leaves
# no density artefact — the cloud above the cutoff is drawn in full, the y-axis simply
# starts at -log10(man.pmax). Benchmarked on 8.4M points: p<=0.2 is 6x faster, p<=0.1 is
# 12x, p<=0.05 is 25x. Set man.pmax <- 1 to draw every SNP.
man.pmax <- 0.1

chr.map <- c(NC_035780.1 = 1, NC_035781.1 = 2, NC_035782.1 = 3, NC_035783.1 = 4,
             NC_035784.1 = 5, NC_035785.1 = 6, NC_035786.1 = 7, NC_035787.1 = 8,
             NC_035788.1 = 9, NC_035789.1 = 10)
treat.lv <- c("CONCON", "STRCON", "CONROD", "STRROD")

# Upstream tools spell these differently (VCF-kit writes TajimaD, not TAJD). Resolve the
# wanted column by name first, then by pattern, so downstream chunks do not care.
resolve.col <- function(d, valcol) {
  if (valcol %in% names(d)) return(valcol)
  pat <- c(TAJD = "taj", PI = "^pi$", OBS_HET = "obs.?het", EXP_HET = "exp.?het")[[valcol]]
  hit <- if (length(pat)) grep(pat, names(d), ignore.case = TRUE, value = TRUE) else character(0)
  if (!length(hit))
    stop("column '", valcol, "' not found; have: ", paste(names(d), collapse = ", "))
  hit[1]
}

# ragg renders large point clouds far faster than cairo; fall back if it is not installed
open.png <- function(file, width, height, res) {
  if (requireNamespace("ragg", quietly = TRUE))
    ragg::agg_png(file, width = width, height = height, units = "in",
                  res = res, background = "transparent")
  else
    png(file, type = "cairo", units = "in", width = width, height = height,
        res = res, bg = "transparent")
}
```

## Gene annotation

```{r}
genes.bed <- fread("./other_files/sorted.ref3.0.gene.bed", header = FALSE, sep = "\t",
                   select = c(1, 2, 3, 10), col.names = c("CHROM", "GSTART", "GEND", "ATTR"))
genes.bed[, GENE := str_match(ATTR, "gene=([^;]+)")[, 2]]

gene.at <- function(cc, pp) {
  g <- genes.bed[CHROM == cc & GSTART <= pp - 1 & GEND > pp - 1, GENE]
  if (!length(g)) "" else paste(unique(g), collapse = "; ")
}

# Readable labels transcribed from the Discussion draft (AZ & CC). NOT machine-verified
# against NCBI — extend as products are confirmed; anything absent falls back to the LOC id.
gene.labels <- c(
  LOC111121590 = "NF-kB p105",        LOC111125856 = "Rho GAP 7",
  LOC111102361 = "Rho GAP 20",        LOC111129890 = "PP2A reg. sub.",
  LOC111127799 = "PI 5-phosphatase",  LOC111102665 = "DBP / HLF",
  LOC111117505 = "Iroquois irx-2",    LOC111129264 = "prospero hox 1",
  LOC111130344 = "Nrf2",              LOC111120511 = "TFIID subunit 2",
  LOC111130101 = "TBP-like",          LOC111107792 = "BRO1 / VGLL4",
  LOC111107402 = "HDAC6",             LOC111137079 = "PTBP3-like",
  LOC111100729 = "metalloendopept.",  LOC111118036 = "cadherin-23",
  LOC111118850 = "titin / SRRM2",     LOC111119752 = "fibrillin-2",
  LOC111122175 = "polycystin-2",      LOC111125411 = "protocadherin b8",
  LOC111108439 = "semaphorin rec.",   LOC111118776 = "Cul2-RING Ub", 
  LOC111124421 = "UNCH/Mucin-like",   LOC111136420 = "glutathione hydrolase-like",
  LOC111099737 = "MsrB-like",         LOC111108426 = "DOCK3-like")


```

## LFMM p-values

The p-value vector is in the same SNP order as `totalloci` (both derive from the biallelic
subset of `SNP.TRSdp10g1.FIL.vcf.gz`), which is the mapping the outlier-index chunks above
already rely on.

```{r}
lfmm.loci <- fread("./analysis/outlier_detection/totalloci",
                   header = FALSE, col.names = c("CHROM", "POS"))

lfmm.pv <- switch(lfmm.var,
  Combined = as.numeric(oys_rod_CB.pv$calibrated.pvalue[, 1]),
  CADO     = as.numeric(adj.pv_rod_CADO),
  Disease  = as.numeric(adj.pv_rod_DIS))
lfmm.qv <- switch(lfmm.var,
  Combined = as.numeric(oys_rod_CB.qv),
  CADO     = as.numeric(oys_rod_CADO.qv),
  Disease  = as.numeric(oys_rod_DIS.qv))

# the join is positional, so a length mismatch means the VCF changed under one of them
stopifnot(nrow(lfmm.loci) == length(lfmm.pv), length(lfmm.pv) == length(lfmm.qv))

lfmm.man <- data.table(CHROM = lfmm.loci$CHROM, POS = lfmm.loci$POS,
                       P = lfmm.pv, Q = lfmm.qv)[CHROM %in% names(chr.map)]

# Some calibrated p-values underflow to exactly 0, which would make -log10(P) infinite and
# blow up the panel A y-scale. Floor them at the smallest non-zero p-value observed, so they
# plot at the top of the real range. P itself is left untouched for ranking and the table.
# If any are floored, say so in the figure legend.
n.zero <- sum(lfmm.man$P <= 0, na.rm = TRUE)
p.floor <- min(lfmm.man$P[lfmm.man$P > 0], na.rm = TRUE)
if (n.zero)
  cat(n.zero, "p-values underflowed to 0; plotted at the smallest non-zero value,",
      format(p.floor, digits = 3), "(-log10 =", round(-log10(p.floor), 1), ")\n")

lfmm.man[, `:=`(CHR = chr.map[CHROM], LOGP = -log10(pmax(P, p.floor)))]
setorder(lfmm.man, CHR, POS)
lfmm.man[, BPcum := seq_len(.N)]
lfmm.axis <- lfmm.man[, .(center = mean(BPcum)), by = CHR]

cat(sum(lfmm.man$Q < 0.1), "SNPs at q < 0.1 for", lfmm.var, "\n")
```

## Candidate regions

The candidate pool for `focal.by = "separation"` is the union outlier set (LFMM ∪ OutFLANK).
The bash chunk computes 500 bp tracks across **all** of these, so scoring and plotting read
the same files and no second pass is needed. 150 loci x 40 kb is ~6 Mb of sequence — seconds.

```{r}
outliers.lfmm <- fread("./analysis/outlier_detection/results/outliers_lfmm2.bed",
                       header = FALSE, col.names = c("CHROM", "START", "POS"))
outliers.of   <- fread("./analysis/FST/Outflank_outliers.bed",
                       header = FALSE, col.names = c("CHROM", "START", "POS"))

cands <- unique(rbind(outliers.lfmm, outliers.of)[, .(CHROM, POS)])[
  CHROM %in% names(chr.map)][order(CHROM, POS)]
cands[, CHR := chr.map[CHROM]]
cands <- lfmm.man[, .(CHROM, POS, P, Q, LOGP, BPcum)][cands, on = .(CHROM, POS)]

dir.create("./analysis/figure2", showWarnings = FALSE)
withr::with_options(c(scipen = 10),
  write.table(data.frame(cands$CHROM, pmax(0, cands$POS - flank - 1), cands$POS + flank),
              "./analysis/figure2/candidate_regions.bed",
              sep = "\t", quote = FALSE, row.names = FALSE, col.names = FALSE))
write(binsize, "./analysis/figure2/binsize.txt")

cat(nrow(cands), "candidate outlier loci\n")
```

## Re-bin diversity metrics across the candidate regions

Same tools and options as the genome-wide sections above, restricted to the candidate
regions and binned at `binsize` instead of 10 kb. **Re-run whenever `flank` or `binsize`
changes.**

```{bash eval=FALSE}
source activate ROD_CADO
cd ./analysis/figure2

VCF=../raw.vcf/filtered/SNP.TRSdp10g1.FIL.vcf.gz
BIN=$(cat binsize.txt)   # set by the `binsize` variable in the R chunk above

sort -k1,1 -k2,2n candidate_regions.bed | bedtools merge -i - > regions.bed

# --- per-SNP FST in the candidate regions (only needed when ins.top = "FST") --
bedtools intersect -a <(sort -k1,1 -k2,2n ../FST/FST_Outflank.bed) \
                   -b regions.bed -wa -sorted \
  | mawk 'BEGIN{print "CHROM\tPOS\tFST\tOUTLIER"} {print $1 "\t" $3 "\t" $4 "\t" $5}' \
  > focal_fst_persnp.txt

# --- nucleotide diversity ------------------------------------------------
for T in con-con str-con con-rod str-rod; do
  bcftools view --threads 20 -R regions.bed $VCF \
      -S ../diversity_statistics/nd/${T}.txt \
    | vcftools --vcf - --window-pi $BIN --out ${T}.focal
done

# --- Tajima's D ----------------------------------------------------------
source activate vkrc
for T in con-con str-con con-rod str-rod; do
  bcftools view --threads 20 -R regions.bed $VCF \
      -S ../diversity_statistics/nd/${T}.txt \
    | vk tajima $BIN $BIN - > ${T}.focal.tajima
done
source activate ROD_CADO

# --- observed heterozygosity --------------------------------------------
# popStats output is already per-site, so this only re-bins the existing *-treat.het files
bedtools makewindows -b regions.bed -w $BIN > focal_bins.bed
for T in CONCON STRCON CONROD STRROD; do
  bedtools intersect -a focal_bins.bed \
                     -b <(sort -k1,1 -k2,2n ../diversity_statistics/het/${T}-treat.het) \
                     -wa -wb \
    | bedtools merge -d -1 -c 7,8 -o mean \
    | mawk -v t=$T '{print $0 "\t" t}' > ${T}.focal.het
done
cat <(echo -e "CHROM\tBIN_START\tBIN_END\tEXP_HET\tOBS_HET\tTREATMENT") \
    CONCON.focal.het STRCON.focal.het CONROD.focal.het STRROD.focal.het > focal.het.txt
```

```{r}
treat.files <- c(CONCON = "con-con", STRCON = "str-con",
                 CONROD = "con-rod", STRROD = "str-rod")

pi.focal <- rbindlist(lapply(names(treat.files), function(t) {
  d <- fread(paste0("./analysis/figure2/", treat.files[t], ".focal.windowed.pi"))
  d$Treatment <- t; d }))

# VCF-kit writes its own header (CHROM / BIN_START / BIN_END / N_Sites / N_SNPs / TajimaD),
# so normalise the column names the way the genome-wide chunk does with its echo -e header.
tajd.focal <- rbindlist(lapply(names(treat.files), function(t) {
  d  <- fread(paste0("./analysis/figure2/", treat.files[t], ".focal.tajima"))
  tj <- grep("taj", names(d), ignore.case = TRUE, value = TRUE)
  setnames(d, if (length(tj)) tj[1] else names(d)[6], "TAJD")
  setnames(d, 1:3, c("CHROM", "BIN_START", "BIN_END"))
  d$Treatment <- t; d }))

het.focal <- fread("./analysis/figure2/focal.het.txt", header = TRUE)
setnames(het.focal, "TREATMENT", "Treatment")

for (d in list(pi.focal, tajd.focal, het.focal))
  d[, Treatment := factor(Treatment, levels = treat.lv)]

stopifnot(all(c("CHROM","BIN_START","BIN_END","Treatment") %in% names(pi.focal)),
          all(c("CHROM","BIN_START","BIN_END","Treatment") %in% names(tajd.focal)),
          all(c("CHROM","BIN_START","BIN_END","Treatment") %in% names(het.focal)))
```

## Focal regions

Focal SNPs within `merge.within` bp of each other are drawn in a single panel: the window is
centred on their midpoint and every member SNP gets its own marker. Adjacent outliers a few
bp apart are the common case — merging them removes a duplicate panel rather than adding
information, so the panel count drops accordingly.

```{r}
if (focal.by == "file") {
  f.path <- "./analysis/figure2/focal_loci.csv"
  if (!file.exists(f.path))
    stop(f.path, " not found — either create it or set focal.by <- \"pvalue\"")
  snps <- fread(f.path)
  stopifnot(all(c("CHROM", "POS") %in% names(snps)))
  if (!"name" %in% names(snps)) {
    snps[, gene := mapply(gene.at, CHROM, POS)]
    snps[, name := ifelse(gene %in% names(gene.labels), gene.labels[gene],
                   ifelse(nzchar(gene), gene, "intergenic"))]
  }
  cat("focal SNPs read from", f.path, "\n")
} else {
  cand2 <- lfmm.man[order(P)]
  keep <- integer(0)
  for (i in seq_len(nrow(cand2))) {
    if (length(keep) >= n.focal) break
    if (length(keep) && any(cand2$CHROM[keep] == cand2$CHROM[i] &
                            abs(cand2$POS[keep] - cand2$POS[i]) < min.sep)) next
    if (genic.only && !nzchar(gene.at(cand2$CHROM[i], cand2$POS[i]))) next
    keep <- c(keep, i)
  }
  snps <- cand2[keep, .(CHROM, POS)]
  snps[, gene := mapply(gene.at, CHROM, POS)]
  snps[, name := ifelse(gene %in% names(gene.labels), gene.labels[gene],
                 ifelse(nzchar(gene), gene, "intergenic"))]
}

## --- group SNPs into panels ----------------------------------------------
setorder(snps, CHROM, POS)
snps[, panel := {
  g <- cumsum(c(TRUE, diff(POS) > merge.within))
  paste(CHROM, g)
}, by = CHROM]
if (merge.within <= 0) snps[, panel := paste(CHROM, POS)]

uniq.or.join <- function(x) paste(unique(x), collapse = " / ")
focal <- snps[, .(CHROM  = CHROM[1],
                  POS    = round(mean(POS)),          # panel centre
                  name   = uniq.or.join(name),
                  theme  = if ("theme" %in% names(snps)) uniq.or.join(na.omit(theme)) else NA_character_,
                  n_snps = .N,
                  score  = if ("sep_score" %in% names(snps)) max(sep_score, na.rm = TRUE) else NA_real_),
              by = panel]

# In "file" mode the list is the choice, so every panel is kept. In "pvalue" mode
# n.focal caps the count, strongest first.
if (!all(is.na(focal$score))) setorder(focal, -score, na.last = TRUE)
focal <- as.data.frame(focal)
if (focal.by != "file") focal <- head(focal, n.focal)
# panels and their labels both run in genome order, so i..n reads left to right in
# panel A and in the grid below it
focal <- focal[order(chr.map[focal$CHROM], focal$POS), ]

focal$CHR <- chr.map[focal$CHROM]
focal$region <- tolower(as.character(as.roman(seq_len(nrow(focal)))))
# 3-row strip label: index+chromosome / gene name / GO theme, stacked instead of
# concatenated onto two lines. Each row is short on its own, so the strip width is
# set by the longest single row rather than the longest combined string.
show.theme <- if (is.na(strip.theme)) TRUE else isTRUE(strip.theme)
focal$theme_row <- if (show.theme)
  ifelse(is.na(focal$theme) | !nzchar(focal$theme), "", focal$theme) else ""

# Word-wrap each row on its own so a single long word/phrase (e.g. "glutathione
# hydrolase-like", "oxidative stress / protein repair") can't run past the panel
# width and get clipped by the strip. strip.wrap is chars/line, not inches \u2014 tune it
# down if columns are still overflowing once there are >7 panels, up if there is room.
strip.wrap <- 14
wrap.row <- function(x, width = strip.wrap)
  vapply(x, function(s) paste(strwrap(s, width = width), collapse = "\n"), character(1))

focal$lab <- paste0(wrap.row(paste0(focal$region, " \u00b7 Chr", focal$CHR)), "\n",
                    wrap.row(focal$name), "\n",
                    wrap.row(focal$theme_row))

# every member SNP, positioned relative to its panel centre — used for the markers
focal.snps <- merge(as.data.table(snps)[, .(panel, CHROM, POS)],
                    as.data.table(focal)[, .(panel, lab, centre = POS)],
                    by = "panel")[, .(region = lab, CHROM, POS, rel = (POS - centre) / 1000)]
focal.snps[, region := factor(region, levels = focal$lab)]

cat(nrow(snps), "focal SNPs ->", nrow(focal), "panels",
    sprintf("(%d merged)", nrow(snps) - nrow(focal)), "\n")
focal[, c("region", "CHROM", "POS", "CHR", "name", "n_snps", "theme")]
```

## Panel A — genome-wide LFMM association

With `man.pmax < 1` the panel is a truncated Manhattan: only SNPs at or below that p-value
are drawn and the y-axis starts at −log10(man.pmax). **Say so in the figure legend.** The
trade-off is that the bulk of the null distribution is no longer visible, so a reader cannot
eyeball inflation from this panel — worth pairing with a QQ plot in the supplement,
especially since the GIF for the Combined variable was 1.41.

```{r}
if (!"BPcum" %in% names(focal))
  focal$BPcum <- vapply(seq_len(nrow(focal)), function(i) {
    sub <- lfmm.man[CHROM == focal$CHROM[i]]
    sub$BPcum[which.min(abs(sub$POS - focal$POS[i]))] }, numeric(1))

# alternate label heights so markers close together on the axis do not overwrite each other
gap <- diff(range(lfmm.man$BPcum)) * 0.03
focal$tier <- 0
if (nrow(focal) > 1)
  for (i in 2:nrow(focal))
    if (focal$BPcum[i] - focal$BPcum[i - 1] < gap)
      focal$tier[i] <- 1 - focal$tier[i - 1]

man.plot <- lfmm.man[P <= man.pmax]
cat(sprintf("panel A: drawing %d of %d SNPs (%.1f%%, p <= %g); y-axis truncated at %.2f\n",
            nrow(man.plot), nrow(lfmm.man), 100 * nrow(man.plot) / nrow(lfmm.man),
            man.pmax, -log10(man.pmax)))

fdr.line  <- if (any(lfmm.man$Q < 0.1)) -log10(max(lfmm.man[Q < 0.1, P])) else NA_real_
lfmm.ymin <- if (man.pmax < 1) -log10(man.pmax) else 0
lfmm.ymax <- max(lfmm.man$LOGP, na.rm = TRUE) * 1.28

fig2.man <- ggplot(man.plot, aes(x = BPcum, y = LOGP)) +
  geom_point(aes(color = as.factor(CHR)), alpha = 0.25, size = 0.4, shape = 16) +
  geom_point(data = lfmm.man[Q < 0.1], aes(color = as.factor(CHR)),
             shape = 17, alpha = 1, size = 1.6) +
  {if (!is.na(fdr.line)) geom_hline(yintercept = fdr.line, linetype = "dashed",
                                    linewidth = 0.3, colour = "grey40")} +
  geom_segment(data = focal, aes(x = BPcum, xend = BPcum,
                                 y = lfmm.ymax * 0.78, yend = lfmm.ymax * (0.88 + tier * 0.055)),
               inherit.aes = FALSE, linewidth = 0.4) +
  geom_text(data = focal, aes(x = BPcum, y = lfmm.ymax * (0.925 + tier * 0.055), label = region),
            inherit.aes = FALSE, size = 3, fontface = "bold") +
  scale_color_manual(values = rep(c("#7d5577", "#52374e"), 5)) +
  scale_x_continuous(label = lfmm.axis$CHR, breaks = lfmm.axis$center,
                     expand = c(0.005, 0)) +
  scale_y_continuous(expand = c(0, 0), limits = c(lfmm.ymin, lfmm.ymax)) +
  guides(colour = "none", alpha = "none", fill = "none") +
  labs(x = "Chromosome", y = "-log<sub>10</sub>(*p*)") +
  theme_classic(base_size = 10) +
  theme(axis.title.y = element_markdown())

if (preview) fig2.man
```

## Optional PCA panel

Ascertainment caveat: the union outlier set includes 107 SNPs that LFMM selected on
`Combined_Bin`, so a PCA of those loci partly recovers the variable they were selected from,
and the PERMANOVA inherits the same circularity. This is why the panel is off by default.

```{r}
# only built when requested, so the figure does not depend on the PCA sections having run
fig2.pca <- NULL
if (show.pca) {
  fig2.pca <- ggplot(pca_data_combined, aes(x = PC1, y = PC2, color = Treatment)) +
    geom_point(size = 2, alpha = 0.85) +
    scale_color_manual(values = col_pal) +
    labs(x = paste0("PC1 (", round(pve_combined[1], 1), "%)"),
         y = paste0("PC2 (", round(pve_combined[2], 1), "%)")) +
    annotate("text", x = -Inf, y = -Inf,
             label = paste("PERMANOVA p =", format(p_value_combined, digits = 2)),
             hjust = -0.15, vjust = -1, size = 3) +
    guides(colour = "none") +
    theme_classic(base_size = 10)
}

if (preview) fig2.pca
```

## Focal region panels

```{r}
top.lab   <- if (ins.top == "LFMM") "-log[10](italic(p))" else "italic(F)[ST]"
metric.lv <- c(top.lab, "pi", "H[o]", "Tajima*minute*s~italic(D)")

top.src <- if (ins.top == "LFMM") {
  lfmm.man[, .(CHROM, POS, VAL = LOGP)]
} else {
  fread("./analysis/figure2/focal_fst_persnp.txt", header = TRUE)[, .(CHROM, POS, VAL = FST)]
}

top.rows <- rbindlist(lapply(seq_len(nrow(focal)), function(i) {
  f <- focal[i, ]
  s <- top.src[CHROM == f$CHROM & abs(POS - f$POS) <= flank][order(POS)]
  if (!nrow(s)) return(NULL)
  data.table(region = f$lab, rel = (s$POS - f$POS) / 1000, value = s$VAL,
             roll = frollmean(s$VAL, min(51, nrow(s)), align = "center"),
             is.focal = s$POS %in% focal.snps[region == f$lab, POS])
}))

grab.focal <- function(d, valcol, metric) {
  d  <- as.data.frame(d)
  vc <- resolve.col(d, valcol)
  rbindlist(lapply(seq_len(nrow(focal)), function(i) {
    f <- focal[i, ]
    s <- d[d$CHROM == f$CHROM &
           d$BIN_START >= f$POS - flank & d$BIN_END <= f$POS + flank, ]
    if (!nrow(s)) return(NULL)
    data.table(region = f$lab,
               rel = ((s$BIN_START + s$BIN_END) / 2 - f$POS) / 1000,
               value = s[[vc]],
               Treatment = as.character(s$Treatment),
               metric = metric)
  }))
}

# centred rolling mean across `smooth.bins` bins, per region x treatment; edges keep raw values
smooth.track <- function(dt, k) {
  if (k <= 1) return(dt)
  dt[order(rel), value := {
    sm <- frollmean(value, k, align = "center", na.rm = TRUE)
    fifelse(is.na(sm), value, sm)
  }, by = .(region, Treatment, metric)]
  dt[]
}

fig2.ins.df <- smooth.track(rbind(
  grab.focal(pi.focal,   "PI",      "pi"),
  grab.focal(het.focal,  "OBS_HET", "H[o]"),
  grab.focal(tajd.focal, "TAJD",    "Tajima*minute*s~italic(D)")), smooth.bins)

fig2.ins.df[, `:=`(metric    = factor(metric, levels = metric.lv),
                   Treatment = factor(Treatment, levels = treat.lv),
                   region    = factor(region, levels = focal$lab))]
top.rows[, `:=`(metric = factor(top.lab, levels = metric.lv),
                region = factor(region, levels = focal$lab))]

# Flag panels built from 2 merged SNPs (the two adjacent-outlier cases) with a "*"
# above their focal marker(s) in the top row, so the merge isn't silently invisible.
# One mark per panel, placed at the higher of the (up to two) focal points — not one
# mark per SNP, so it reads as "this panel = 2 loci" rather than a per-point flag.
merged.lab <- focal$lab[focal$n_snps > 1]
top.marks <- top.rows[is.focal == TRUE & region %in% merged.lab,
                      .(rel = mean(rel), value = max(value, na.rm = TRUE)), by = region]
if (nrow(top.marks)) top.marks[, metric := factor(top.lab, levels = metric.lv)]

fig2.ins <- ggplot(fig2.ins.df, aes(x = rel, y = value, colour = Treatment)) +
  geom_vline(data = focal.snps, aes(xintercept = rel),
             linetype = "dashed", linewidth = 0.3, colour = "grey55") +
  geom_point(data = top.rows, inherit.aes = FALSE, aes(rel, value),
             colour = "grey78", size = 0.2, alpha = 0.45, shape = 16) +
  geom_line(data = top.rows, inherit.aes = FALSE, aes(rel, roll),
            colour = "grey20", linewidth = 0.5, na.rm = TRUE) +
  geom_point(data = top.rows[is.focal == TRUE], inherit.aes = FALSE, aes(rel, value),
             shape = 17, size = 1.9, colour = "#B22222") +
  {if (nrow(top.marks)) geom_text(data = top.marks, inherit.aes = FALSE,
             aes(rel, value, label = "*"), vjust = 0, size = 4.5,
             fontface = "bold", colour = "#B22222")} +
  geom_line(linewidth = 0.45, na.rm = TRUE) +
  facet_grid(metric ~ region, scales = "free_y", strip.clip = "off",
             labeller = labeller(metric = label_parsed)) +
  scale_colour_manual(values = col_pal, name = "Treatment") +
  scale_x_continuous(breaks = c(-10, 0, 10)) +
  labs(x = "Distance from focal outlier SNP (kb)", y = NULL) +
  theme_classic(base_size = 9) +
  theme(strip.background = element_blank(),
        strip.text.x = element_text(size = 6.5, face = "bold", lineheight = 1.1),
        strip.text.y = element_text(size = 8),
        panel.spacing = unit(3, "pt"))

if (preview) fig2.ins
```

## Assemble

```{r}
fig2 <- if (show.pca) {
  fig2.man / (fig2.pca | fig2.ins) +
    plot_layout(heights = c(1, 2.6), widths = c(0.62, 1.38), guides = "collect")
} else {
  fig2.man / fig2.ins + plot_layout(heights = c(1, 2.4), guides = "collect")
}

fig2 <- fig2 + plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom",
        plot.tag = element_text(face = "bold", size = 12))

if (preview) fig2
```

```{r}
fig.w <- if (show.pca) 7.5 else max(7.5, 1.55 * nrow(focal))
fig.h <- if (show.pca) 9 else 8

open.png("figures/Figure2.png", width = fig.w, height = fig.h, res = 300)
print(fig2)
null <- dev.off()
```

# Gene Ontology Enrichment Analysis — Combined Outliers (LFMM2 ∪ OutFLANK)
Original analysis by Amy Zyck, edits by JBP

## Identifying genes for candidate loci

Using the combined outlier BED file (LFMM2 + OutFLANK) generated above, we intersect with the oyster genome gene annotation to identify candidate genes.

```{bash eval=FALSE}
source activate ROD_CADO
mkdir -p ./analysis/GO
cd ./analysis/GO

# Use the combined outlier BED already generated
bedtools intersect -wb -a ../outlier_detection/results/outliers_lfmm2_outflank_combined.bed -b ../../other_files/sorted.ref3.0.gene.bed | grep -oh "gene=LOC.*;g" | sed 's/gene=//g' | sed 's/;g//g' | sort | uniq > Sig.gene.ROD.CADO.LOC

cat Sig.gene.ROD.CADO.LOC | wc -l
```

## Gene Ontology with TopGO

### Download and prepare GO annotation

```{bash eval=FALSE}
source activate ROD_CADO
cd ./analysis/GO

wget https://ftp.ncbi.nlm.nih.gov/genomes/all/GCF/002/022/765/GCF_002022765.2_C_virginica-3.0/GCF_002022765.2_C_virginica-3.0_gene_ontology.gaf.gz

gunzip GCF_002022765.2_C_virginica-3.0_gene_ontology.gaf.gz

mv GCF_002022765.2_C_virginica-3.0_gene_ontology.gaf gene_ontology.gaf
```

### Create gene-to-GO mapping file

Making a tab file with gene ID and GO IDs. Each gene has one row with all associated GO terms separated by semicolons.

```{bash eval=FALSE}
cd ./analysis/GO

awk 'NR>9 {print $3 "\t" $5}' gene_ontology.gaf | 
sed '1i gene_name\tGeneOntologyIDs' | 
sort | awk '
NR==1 {print; next}
$1!=prev {
if (NR>2) print prev "\t" goids;
prev=$1;
goids=$2;
next
}
{
goids=goids "; " $2
}
END {print prev "\t" goids}
' > gene2go_full.tab
```

### Extract GO terms for candidate genes

```{bash eval=FALSE}
cd ./analysis/GO

grep -F -f Sig.gene.ROD.CADO.LOC gene2go_full.tab > ROD_cand_genes_GO.tab

# Add header
awk 'BEGIN {print "gene_name\tGeneOntologyIDs"} {print}' ROD_cand_genes_GO.tab > ROD_cand_genes_GO_head.tab
```

### Load gene-to-GO data
```{r}
all_gene2go <- read.delim("analysis/GO/gene2go_full.tab", sep="\t")
out_gene2go <- read.delim("analysis/GO/ROD_cand_genes_GO_head.tab", sep="\t")
```

### Prepare gene list for TopGO
```{r}
cand_genes <- as.character(unique(out_gene2go$gene_name))
all_genes <- as.character(unique(all_gene2go$gene_name))
out_GeneList <- factor(as.integer(all_genes %in% cand_genes))
names(out_GeneList) <- all_genes
```

### Read in gene-to-GO mappings
```{r}
gene2go_topgo <- readMappings("analysis/GO/gene2go_full.tab", IDsep=";", sep="\t")
```

### Set gene selection function
```{r}
topDiffGenes <- function(allScore) {
  return(allScore < 0.05)
}
```

### Biological Processes

```{r}
GO_OUT_BP <- new("topGOdata", ontology="BP", gene2GO=gene2go_topgo, allGenes=out_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)
```

```{r}
GO_OUT_BP_FE <- runTest(GO_OUT_BP, algorithm="weight01", statistic="fisher")
```

```{r}
GO_OUT_BP_En <- GenTable(GO_OUT_BP, Fisher = GO_OUT_BP_FE, orderBy = "Fisher", topNodes = 100, numChar = 100)
```

```{r}
GO_OUT_BP_En$Fisher <- as.numeric(GO_OUT_BP_En$Fisher)
GO_OUT_BP_En_sig <- GO_OUT_BP_En[GO_OUT_BP_En$Fisher < 0.05,]
```

```{r}
# Separate GO terms
out_gene2go_sep <- out_gene2go %>%
  separate_rows(GeneOntologyIDs, sep = ";")

out_gene2go_sep$GeneOntologyIDs <- trimws(out_gene2go_sep$GeneOntologyIDs)
GO_OUT_BP_En_sig$GO.ID <- trimws(GO_OUT_BP_En_sig$GO.ID)

GO_OUT_BP_En_sig_gene <- out_gene2go_sep %>%
  left_join(GO_OUT_BP_En_sig, by = c("GeneOntologyIDs" = "GO.ID")) %>%
  na.omit()

GO_OUT_BP_En_sig_gene$ontology <- "Biological Processes"
```

### Cellular Components

```{r}
GO_OUT_CC <- new("topGOdata", ontology="CC", gene2GO=gene2go_topgo, allGenes=out_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)
```

```{r}
GO_OUT_CC_FE <- runTest(GO_OUT_CC, algorithm="weight01", statistic="fisher")
```

```{r}
GO_OUT_CC_En <- GenTable(GO_OUT_CC, Fisher = GO_OUT_CC_FE, orderBy = "Fisher", topNodes = 100, numChar = 100)
```

```{r}
GO_OUT_CC_En$Fisher <- as.numeric(GO_OUT_CC_En$Fisher)
GO_OUT_CC_En_sig <- GO_OUT_CC_En[GO_OUT_CC_En$Fisher < 0.05,]
```

```{r}
GO_OUT_CC_En_sig$GO.ID <- trimws(GO_OUT_CC_En_sig$GO.ID)

GO_OUT_CC_En_sig_gene <- out_gene2go_sep %>%
  left_join(GO_OUT_CC_En_sig, by = c("GeneOntologyIDs" = "GO.ID")) %>%
  na.omit()

GO_OUT_CC_En_sig_gene$ontology <- "Cellular Components"
```

### Molecular Functions

```{r}
GO_OUT_MF <- new("topGOdata", ontology="MF", gene2GO=gene2go_topgo, allGenes=out_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)
```

```{r}
GO_OUT_MF_FE <- runTest(GO_OUT_MF, algorithm="weight01", statistic="fisher")
```

```{r}
GO_OUT_MF_En <- GenTable(GO_OUT_MF, Fisher = GO_OUT_MF_FE, orderBy = "Fisher", topNodes = 100, numChar = 100)
```

```{r}
GO_OUT_MF_En$Fisher <- as.numeric(GO_OUT_MF_En$Fisher)
GO_OUT_MF_En_sig <- GO_OUT_MF_En[GO_OUT_MF_En$Fisher < 0.05,]
```

```{r}
GO_OUT_MF_En_sig$GO.ID <- trimws(GO_OUT_MF_En_sig$GO.ID)

GO_OUT_MF_En_sig_gene <- out_gene2go_sep %>%
  left_join(GO_OUT_MF_En_sig, by = c("GeneOntologyIDs" = "GO.ID")) %>%
  na.omit()

GO_OUT_MF_En_sig_gene$ontology <- "Molecular Functions"
```

### Combine all ontologies

```{r}
GO_OUT_En_sig_gene <- rbind(GO_OUT_BP_En_sig_gene, GO_OUT_CC_En_sig_gene, GO_OUT_MF_En_sig_gene)
```

```{r}
merged_out_genes <- GO_OUT_En_sig_gene %>%
  mutate(prop.sig.genes = Significant/Annotated)

# Save unique instances of gene/GO pairs
merged_out_genes <- unique(merged_out_genes)
```

```{r}
write.csv(merged_out_genes, "analysis/GO/ROD_OUT_GO_en_sig_gene.csv")
```

### Plot GO enrichment

```{r}
GO_plot <- ggplot(merged_out_genes, aes(x = Term, y = -log(Fisher), size = prop.sig.genes, fill = -log(Fisher))) +
  expand_limits(y = 1.5) +
  ylim(1, 7.25) +
  geom_hline(yintercept = -log10(0.01), linetype = "longdash", colour = "black", linewidth = .6) +
  geom_hline(yintercept = -log10(0.001), linetype = "solid", colour = "black", linewidth = .6) +
  geom_point(shape = 21) + 
  scale_size(range = c(2, 12)) + 
  scale_fill_continuous(low = "#1AD3D1FF", high = "#4686FBFF") +
  xlab('') + 
  ylab('Enrichment score') +
  theme_bw() +
  facet_grid(vars(ontology), scales = "free", space = "free_y") +
  coord_flip()

GO_plot
```

```{r}
png(filename = "figures/GO_enrichment_plot.png",
    type = "cairo",
    width = 2000,
    height = 2000,
    res = 200,
    bg = "transparent")

print(GO_plot)
dev.off()
```

# Gene Ontology Enrichment Analysis — Overlapping Outliers Only (LFMM2 ∩ OutFLANK)
Original analysis by Amy Zyck, edits by JBP

This section mirrors the GO enrichment analysis above but restricts candidate loci to SNPs identified as outliers by **both** LFMM2 and OutFLANK (i.e., the intersection). This allows a direct comparison of enriched terms between the union set and the high-confidence overlap set.

## Identifying genes for overlapping candidate loci

```{bash eval=FALSE}
source activate ROD_CADO
cd ./analysis/GO

# Use the overlapping outlier BED already generated above
bedtools intersect -wb -a ../outlier_detection/results/outliers_overlap_lfmm2_outflank.bed -b ../../other_files/sorted.ref3.0.gene.bed | grep -oh "gene=LOC.*;g" | sed 's/gene=//g' | sed 's/;g//g' | sort | uniq > Sig.gene.ROD.CADO.overlap.LOC

cat Sig.gene.ROD.CADO.overlap.LOC | wc -l
```

### Extract GO terms for overlapping candidate genes

```{bash eval=FALSE}
cd ./analysis/GO

grep -F -f Sig.gene.ROD.CADO.overlap.LOC gene2go_full.tab > ROD_overlap_genes_GO.tab

# Add header
awk 'BEGIN {print "gene_name\tGeneOntologyIDs"} {print}' ROD_overlap_genes_GO.tab > ROD_overlap_genes_GO_head.tab
```

### Load gene-to-GO data for overlapping outliers
```{r}
overlap_gene2go <- read.delim("analysis/GO/ROD_overlap_genes_GO_head.tab", sep="\t")
```

### Prepare gene list for TopGO
```{r}
overlap_cand_genes <- as.character(unique(overlap_gene2go$gene_name))
overlap_GeneList <- factor(as.integer(all_genes %in% overlap_cand_genes))
names(overlap_GeneList) <- all_genes
```

### Biological Processes

```{r}
GO_OV_BP <- new("topGOdata", ontology="BP", gene2GO=gene2go_topgo, allGenes=overlap_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)
GO_OV_BP_FE <- runTest(GO_OV_BP, algorithm="weight01", statistic="fisher")
GO_OV_BP_En <- GenTable(GO_OV_BP, Fisher = GO_OV_BP_FE, orderBy = "Fisher", topNodes = 100, numChar = 100)
GO_OV_BP_En$Fisher <- as.numeric(GO_OV_BP_En$Fisher)
GO_OV_BP_En_sig <- GO_OV_BP_En[GO_OV_BP_En$Fisher < 0.05,]
```

```{r}
overlap_gene2go_sep <- overlap_gene2go %>%
  separate_rows(GeneOntologyIDs, sep = ";")
overlap_gene2go_sep$GeneOntologyIDs <- trimws(overlap_gene2go_sep$GeneOntologyIDs)
GO_OV_BP_En_sig$GO.ID <- trimws(GO_OV_BP_En_sig$GO.ID)

GO_OV_BP_En_sig_gene <- overlap_gene2go_sep %>%
  left_join(GO_OV_BP_En_sig, by = c("GeneOntologyIDs" = "GO.ID")) %>%
  na.omit()
GO_OV_BP_En_sig_gene$ontology <- "Biological Processes"
```

### Cellular Components

```{r}
GO_OV_CC <- new("topGOdata", ontology="CC", gene2GO=gene2go_topgo, allGenes=overlap_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)
GO_OV_CC_FE <- runTest(GO_OV_CC, algorithm="weight01", statistic="fisher")
GO_OV_CC_En <- GenTable(GO_OV_CC, Fisher = GO_OV_CC_FE, orderBy = "Fisher", topNodes = 100, numChar = 100)
GO_OV_CC_En$Fisher <- as.numeric(GO_OV_CC_En$Fisher)
GO_OV_CC_En_sig <- GO_OV_CC_En[GO_OV_CC_En$Fisher < 0.05,]
```

```{r}
GO_OV_CC_En_sig$GO.ID <- trimws(GO_OV_CC_En_sig$GO.ID)

GO_OV_CC_En_sig_gene <- overlap_gene2go_sep %>%
  left_join(GO_OV_CC_En_sig, by = c("GeneOntologyIDs" = "GO.ID")) %>%
  na.omit()
GO_OV_CC_En_sig_gene$ontology <- "Cellular Components"
```

### Molecular Functions

```{r}
GO_OV_MF <- new("topGOdata", ontology="MF", gene2GO=gene2go_topgo, allGenes=overlap_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)
GO_OV_MF_FE <- runTest(GO_OV_MF, algorithm="weight01", statistic="fisher")
GO_OV_MF_En <- GenTable(GO_OV_MF, Fisher = GO_OV_MF_FE, orderBy = "Fisher", topNodes = 100, numChar = 100)
GO_OV_MF_En$Fisher <- as.numeric(GO_OV_MF_En$Fisher)
GO_OV_MF_En_sig <- GO_OV_MF_En[GO_OV_MF_En$Fisher < 0.05,]
```

```{r}
GO_OV_MF_En_sig$GO.ID <- trimws(GO_OV_MF_En_sig$GO.ID)

GO_OV_MF_En_sig_gene <- overlap_gene2go_sep %>%
  left_join(GO_OV_MF_En_sig, by = c("GeneOntologyIDs" = "GO.ID")) %>%
  na.omit()
GO_OV_MF_En_sig_gene$ontology <- "Molecular Functions"
```

### Combine all ontologies

```{r}
GO_OV_En_sig_gene <- rbind(GO_OV_BP_En_sig_gene, GO_OV_CC_En_sig_gene, GO_OV_MF_En_sig_gene)

merged_overlap_genes <- GO_OV_En_sig_gene %>%
  mutate(prop.sig.genes = Significant / Annotated)
merged_overlap_genes <- unique(merged_overlap_genes)

write.csv(merged_overlap_genes, "analysis/GO/ROD_OVERLAP_GO_en_sig_gene.csv")
```

### Plot GO enrichment — overlapping outliers only

```{r}
GO_plot_overlap <- ggplot(merged_overlap_genes, aes(x = Term, y = -log(Fisher), size = prop.sig.genes, fill = -log(Fisher))) +
  expand_limits(y = 1.5) +
  ylim(1, 7.25) +
  geom_hline(yintercept = -log10(0.01), linetype = "longdash", colour = "black", linewidth = .6) +
  geom_hline(yintercept = -log10(0.001), linetype = "solid", colour = "black", linewidth = .6) +
  geom_point(shape = 21) +
  scale_size(range = c(2, 12)) +
  scale_fill_continuous(low = "#1AD3D1FF", high = "#4686FBFF") +
  xlab('') +
  ylab('Enrichment score') +
  theme_bw() +
  facet_grid(vars(ontology), scales = "free", space = "free_y") +
  coord_flip()

GO_plot_overlap
```

```{r}
png(filename = "figures/GO_enrichment_overlap_plot.png",
    type = "cairo",
    width = 2000,
    height = 2000,
    res = 200,
    bg = "transparent")
print(GO_plot_overlap)
dev.off()
```

## Side-by-side comparison: Combined vs. Overlapping outliers

To directly compare enriched GO terms between the two outlier sets, we add a grouping label to each dataset and bind them together.

```{r}
merged_out_genes$outlier_set    <- "Combined (LFMM2 \u222a OutFLANK)"
merged_overlap_genes$outlier_set <- "Overlap (LFMM2 \u2229 OutFLANK)"

GO_comparison <- bind_rows(merged_out_genes, merged_overlap_genes)
```

```{r}
GO_comparison_plot <- ggplot(GO_comparison, aes(x = Term, y = -log(Fisher), size = prop.sig.genes, fill = -log(Fisher))) +
  expand_limits(y = 1.5) +
  ylim(1, 7.25) +
  geom_hline(yintercept = -log10(0.01), linetype = "longdash", colour = "black", linewidth = .6) +
  geom_hline(yintercept = -log10(0.001), linetype = "solid", colour = "black", linewidth = .6) +
  geom_point(shape = 21) +
  scale_size(range = c(2, 12)) +
  scale_fill_continuous(low = "#1AD3D1FF", high = "#4686FBFF") +
  xlab('') +
  ylab('Enrichment score') +
  theme_bw() +
  facet_grid(vars(ontology, outlier_set), scales = "free", space = "free_y") +
  coord_flip() +
  theme(strip.text.y = element_text(angle = 0, size = 8))

GO_comparison_plot
```

```{r}
png(filename = "figures/GO_enrichment_comparison_plot.png",
    type = "cairo",
    width = 3000,
    height = 2500,
    res = 200,
    bg = "transparent")
print(GO_comparison_plot)
dev.off()
```
